---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.width = 8,
  fig.height = 5
)
```

# ackwards

[![R-CMD-check](https://github.com/jmgirard/ackwards/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jmgirard/ackwards/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/jmgirard/ackwards/branch/master/graph/badge.svg)](https://app.codecov.io/gh/jmgirard/ackwards?branch=master)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)

> Bass-ackwards hierarchical structural analysis in R

**ackwards** implements Goldberg's (2006) bass-ackwards method and its modern
extensions for mapping the hierarchical structure of multivariate data.

The core insight is simple: instead of asking *how many* factors best describe
your data, ask *how the solutions at different levels of resolution are related
to one another*. A 1-factor solution captures the broadest shared variance; a
5-factor solution captures narrower, more specific dimensions. Bass-ackwards
analysis traces how broad factors split into narrow ones — and which narrow
factors are redundant re-combinations of broader ones.

The package supports three extraction engines (PCA, EFA, ESEM), polychoric
correlations for ordinal data, and Forbes's (2023) extension that reveals
skip-level connections and flags redundant or artifactual factors.

## Installation

```{r install, eval = FALSE}
# install.packages("pak")
pak::pak("jmgirard/ackwards")
```

## Quick start

We use `bfi25`, the built-in 25-item Big Five example dataset (see `?bfi25` for
provenance). Because BFI items are recorded on a 6-point ordinal scale, we set
`cor = "polychoric"`.

### Step 1 — Suggest a range of k

`suggest_k()` runs five complementary criteria — two forms of parallel analysis
(PC and FA basis), MAP, VSS, and optionally Comparison Data — to help you choose
an upper bound for the hierarchy depth.

```{r suggest}
library(ackwards)

bfi <- na.omit(bfi25)
sk <- suggest_k(bfi)
sk
```

The criteria converge on a consensus range that covers k = 5, consistent with
the known Big Five structure of this instrument.

### Step 2 — Fit the hierarchy

`ackwards()` fits factor models at every level from 1 to `k_max` and computes the
between-level factor-score correlations that define the hierarchy.

```{r fit}
x <- ackwards(bfi, k_max = 5, cor = "polychoric")
x
```

### Step 3 — Visualize

`autoplot()` draws the hierarchical diagram. Each row is a level (k = 1 at top,
k = 5 at bottom); arrows show which narrow factors inherit from which broad
factor. Arrow **thickness** encodes |r| and arrow **color** encodes direction
(blue = positive, red = negative), each with its own legend. Primary-parent
edges are always positive after sign alignment, so a red arrow marks a genuine
secondary (cross-branch) relationship. Pass `direction = "horizontal"` for a
left-to-right layout.

```{r plot, message = FALSE, warning = FALSE, dev = "png", dpi = 150}
autoplot(x)
```

The five-factor level cleanly splits into the Big Five. The single broad factor
at k = 1 (roughly *general positive character*) differentiates first into
positive vs. negative affect (k = 2), then into successively narrower traits.

### Step 4 — Inspect and score

`tidy()` extracts any part of the object into a tidy data frame; `augment()`
appends factor scores to your data for downstream analysis.

```{r tidy}
# Eight strongest adjacent-level primary edges
edges <- tidy(x, what = "edges", sort = "strength")
head(edges[edges$is_primary, c("from", "to", "r")], 8)
```

```{r augment}
# Append scores for all 5 levels to the original data frame
scored <- augment(x, data = bfi)
dim(scored) # original 25 items + 1+2+3+4+5 = 15 score columns
names(scored)[26:40]
```

## Learn more

| Vignette | Topic |
|----------|-------|
| [Introduction](https://jmgirard.github.io/ackwards/articles/ackwards-intro.html) | Full PCA walkthrough: `suggest_k` → `ackwards` → inspect → plot → score |
| [Choosing k](https://jmgirard.github.io/ackwards/articles/ackwards-suggest-k.html) | Five criteria explained: pros/cons, bias direction, engine pairing |
| [Engines & rotation](https://jmgirard.github.io/ackwards/articles/ackwards-engines.html) | When to choose EFA or ESEM over PCA; convergence and loading comparison |
| [Ordinal data](https://jmgirard.github.io/ackwards/articles/ackwards-ordinal.html) | Polychoric correlations, attenuation bias, and WLSMV estimation |
| [Forbes extension](https://jmgirard.github.io/ackwards/articles/ackwards-forbes.html) | Skip-level edges, redundancy pruning, `pairs = "all"` |
| [Visualization](https://jmgirard.github.io/ackwards/articles/ackwards-visualization.html) | Styling `autoplot()`: sign/magnitude encoding, layout orientation, labels, publication figures |
| [Interpreting & labeling](https://jmgirard.github.io/ackwards/articles/ackwards-interpret.html) | `top_items()`, hierarchy-aware naming, `label_template()` round-trip |

## Citation

If you use **ackwards** in your research, please cite the package:

```{r citation}
citation("ackwards")
```

Please also cite the relevant method paper(s): Goldberg (2006)
<https://doi.org/10.1016/j.jrp.2006.01.001> for the bass-ackwards method itself;
Waller (2007) <https://doi.org/10.1016/j.jrp.2006.08.005> if you rely on the
exact `W'RW` edge algebra; and Forbes (2023)
<https://doi.org/10.1037/met0000546> if you use the extended method
(`pairs = "all"`, redundancy/artifact pruning).
