---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ackwards

<!-- badges: start -->
<!-- badges: end -->

The goal of {ackwards} is to provide functions for the traditional and extended bass-ackwards techniques.

## Installation

You can install the development version of {ackwards} from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("jmgirard/ackwards")
```

## Example

Work in progress on the extended bass-ackwards technique:

```{r example}
library(ackwards)
xba_psych(r = forbes, nfactors = 10, fm = "minres")
```

## References

- Goldberg, L. R. (2006). Doing it all bass-ackwards: The development of hierarchical factor structures from the top down. *Journal of Research in Personality, 40*(4), 347–358. <https://doi.org/10/c9sqkd>

- Waller, N. (2007). A general method for computing hierarchical component structures by Goldberg's bass-ackwards method. *Journal of Research in Personality, 41*(4), 745–752. <https://doi.org/10/bcz8wd>

- Forbes, M. K. (2022). Improving hierarchical models of individual differences: An extension of Goldberg's bass-ackwards method. *Psychological Methods.* Advanced online publication. <https://doi.org/10.1037/met0000546>
