---
title: "ackwards: An R Package for Bass-Ackwards Hierarchical Structural Analysis"
shorttitle: "The ackwards Package"
author:
  - name: Jeffrey M. Girard
    corresponding: true
    orcid: 0000-0002-7359-3746
    email: me@jmgirard.com
    affiliations:
      - name: University of Kansas
        department: Department of Psychology
        city: Lawrence
        region: KS
        country: USA
author-note:
  disclosures:
    conflict of interest: The author has no conflicts of interest to disclose.
    gratitude: |
      During the preparation of this work the author used a large language model
      (Anthropic's Claude, via the Claude Code interface) to assist with software
      implementation, testing, documentation, and editing; the author reviewed
      and verified all output and takes full responsibility for the content. The
      AI system is not an author. See the Use of Generative AI section for
      details.
abstract: |
  Factor analysis confronts every investigator with the same question — how many
  factors to retain — yet solutions at different levels of resolution are
  complementary rather than competing. Goldberg's (2006) bass-ackwards method
  makes this explicit by extracting factor solutions from one to *k* factors and
  characterizing the resulting hierarchy through between-level factor-score
  correlations. We present **ackwards**, an R package implementing the
  bass-ackwards method and its modern descendants across principal component
  analysis, exploratory factor analysis, and exploratory structural equation
  modeling engines, together with Forbes's (2023) all-levels and
  redundancy-pruning extensions. The package computes between-level correlations
  through Waller's (2007) exact closed-form algebra, keeps a light and shareable
  result object, and exposes a tidy, pipe-friendly interface for fitting,
  scoring, visualizing, and pruning a hierarchy. We illustrate the workflow on a
  Big Five personality dataset and reproduce Forbes's 155-variable applied
  example, whose between-level correlations match her reference implementation to
  machine precision.
keywords: [factor analysis, hierarchical structure, bass-ackwards, R package, psychometrics]
bibliography: references.bib
format:
  apaquarto-pdf: default
  apaquarto-docx: default
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(ackwards)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 300
)
options(digits = 3)
```

<!-- ======================================================================= -->
<!-- INTRODUCTION — author-owned stub. Draft the scholarly framing here.     -->
<!-- ======================================================================= -->

# Introduction

> **\[AUTHOR TO DRAFT — structured stub]**
> The seeded Method and Package sections below supply the technical substance;
> the Introduction should supply the scholarly framing and argument. Suggested
> beats:
>
> - The recurring "how many factors?" problem and why competing single-solution
>   answers misframe it [@goldberg2006].
> - The appeal of a *hierarchical* view of individual-differences structure
>   (personality, psychopathology) and the empirical questions it answers.
> - The gap: bass-ackwards analyses have been run with bespoke scripts
>   [@goldberg2006; @waller2007; @forbes2023], with no general, tested,
>   documented tool — motivating a package.
> - A one-paragraph roadmap of the paper.

<!-- ======================================================================= -->
<!-- THE BASS-ACKWARDS METHOD — seeded draft (Method background).             -->
<!-- ======================================================================= -->

# The bass-ackwards method

Anyone who applies factor analysis faces the same dilemma: how many factors
should be extracted? Parallel analysis might indicate five, a scree plot three,
and a reviewer one. It is tempting to treat these as competing answers to a
single question. @goldberg2006 reframed the problem: the solutions are not
competing but complementary. A one-factor solution captures the broadest shared
variance in the data — the dimension along which all variables correlate — while
a five-factor solution captures narrower, more specific dimensions. The two
levels describe the same underlying structure from different vantage points, much
as a satellite image and a street map describe the same city.

The bass-ackwards method makes this relationship explicit. It fits factor models
at every level from 1 up to *k* and then computes the correlations between factor
*scores* across levels. Those correlations reveal which narrow factor inherits
from which broad factor (the lineage of each dimension as resolution increases),
where a broad factor *splits* into sub-dimensions, and where a factor is *stable*
— correlating near 1.0 with its counterpart at an adjacent level, and thus
essentially the same construct at two resolutions. The result is a hierarchy map:
a visual and numerical account of how the factor structure builds from broad to
narrow. Importantly, this is a descriptive, data-driven hierarchy, not a
confirmatory model such as Schmid–Leiman or higher-order structural equation
modeling; the between-level correlations are score correlations (or their
algebraic equivalents), not model parameters.

## Computing between-level correlations

The between-level correlations are the heart of the method, and there are two
routes to them. The direct route materializes factor scores at each level and
correlates them. @waller2007 showed that, when scoring is a linear function of
the items, the same correlations follow in closed form from the factor weight
matrices and the item correlation matrix, without ever computing scores. For
factor weights $W_a$ and $W_b$ at two levels and item correlation matrix $R$, the
cross-level score covariance is $W_a' R W_b$; standardizing by the score standard
deviations $\sqrt{\operatorname{diag}(W'RW)}$ yields the correlations. **ackwards**
uses this algebra by default and falls back to materialized scores only for
nonlinear scoring, cross-checking the two routes against one another for linear
engines. Orthogonal (varimax) rotation within each level is what makes the
algebra exact: an orthonormal rotation satisfies $T' = T^{-1}$, so the closed
form holds rather than approximates. Oblique rotation is deliberately not
offered, because correlated factors would confound the very cross-level signal
the method exists to measure.

## The Forbes extension

@goldberg2006 computed correlations only between *adjacent* levels (1↔2, 2↔3, …).
@forbes2023 extended the method in two ways. First, correlations are computed
across *all* level pairs, not just neighbors, which exposes skip-level
connections: a factor that persists nearly unchanged across several levels
correlates near 1.0 with its distant counterpart, revealing that the intermediate
levels add refinement rather than new information. Second, these all-pairs
correlations support automatic detection of **redundant** factors — chains of
near-identical factors across levels — and researcher-guided inspection of
**artifactual** factors, whose loading patterns duplicate a factor elsewhere in
the hierarchy (measured by Tucker's congruence coefficient; @lorenzoseva2006) or
whose structure appears under-identified. Forbes's redundancy chase follows the
*direct* correlation from a factor to each ancestor level, the criterion
**ackwards** adopts by default.

<!-- ======================================================================= -->
<!-- THE ACKWARDS PACKAGE — seeded draft (Statement of Need + design).        -->
<!-- ======================================================================= -->

# The ackwards package

Despite three decades of use, bass-ackwards analyses have been conducted with
bespoke, paper-specific scripts [@goldberg2006; @waller2007; @forbes2023]. No
general implementation existed that (a) spans the extraction engines researchers
actually use, (b) computes between-level correlations through a single, verified
path, (c) reproduces the published method exactly, and (d) is documented and
tested to the standard expected of research software. **ackwards** fills that
gap. It is available from CRAN and developed openly at
`https://github.com/jmgirard/ackwards`.

The package builds a hierarchy with a single call, `ackwards(data, k_max, ...)`,
dispatching to one of three extraction engines behind a common interface:
principal component analysis (`"pca"`, the default), exploratory factor analysis
(`"efa"`), and exploratory structural equation modeling (`"esem"`, an
EFA-in-SEM formulation via **lavaan** [@lavaan] that adds per-level fit indices
and standard errors). Component and factor extraction rest on the **psych**
package [@psych]. Correlation input may be Pearson (default), Spearman, or
polychoric for ordinal items, and a consequential auto-choice — the basis, the
number of factors, an estimator — is always announced rather than made silently.
Non-convergence at a deep level is treated as information, not failure: the level
is recorded, warned about, and skipped, and the hierarchy is still built to the
deepest converged level.

The result is a light object carrying loadings, variance, fit, weights, edges,
lineage, and the input correlation matrix; heavy artifacts such as factor scores
and raw fits are optional and recomputable, keeping results small and shareable.
A tidy, pipe-friendly surface supports the full workflow: `suggest_k()` for a
plausible range of hierarchy depths, `ackwards()` to fit, `print()` / `summary()`
/ `glance()` to inspect, `autoplot()` to draw the hierarchy diagram, `tidy()` and
`top_items()` to extract and read factors, `augment()` / `predict()` to score
observations in or out of sample, and `prune()` to apply the Forbes redundancy
and artifact diagnostics. The two sections that follow demonstrate this workflow:
first a compact illustrative walkthrough, then Forbes's full applied example.

<!-- ======================================================================= -->
<!-- ILLUSTRATIVE WALKTHROUGH — live from the package (bfi25).                -->
<!-- ======================================================================= -->

# An illustrative walkthrough

We use `bfi25`, a built-in dataset of 1,000 respondents answering 25 personality
items measuring the Big Five traits — Agreeableness, Conscientiousness,
Extraversion, Neuroticism, and Openness — each on a six-point scale. For this
illustration we restrict to respondents with complete data:

```{r}
#| label: data
bfi <- na.omit(bfi25)
dim(bfi)
```

Before fitting, `suggest_k()` reports several complementary factor-retention
criteria [@horn1965; @velicer1976] and a consensus range, rather than a single
number:

```{r}
#| label: suggest-k
sk <- suggest_k(bfi, seed = 42)
sk
```

Because the items are ordinal, we fit on the polychoric correlation basis. A
single call builds the hierarchy from one to five factors:

```{r}
#| label: fit
x <- ackwards(bfi, k_max = 5, cor = "polychoric")
```

The hierarchy diagram (@fig-bfi) is the centerpiece of the method. Each row is a
level (*k* = 1 at the top), and each arrow connects a factor to its *primary
parent* — the factor one level up with which it correlates most strongly. Arrow
thickness is proportional to $|r|$ and color encodes sign.

```{r}
#| label: fig-bfi
#| fig-cap: "Bass-ackwards hierarchy of the 25 Big Five items, extracted from one to five factors on a polychoric basis. Arrows connect each factor to its primary parent; thickness is proportional to the absolute between-level score correlation."
#| fig-width: 8
#| fig-height: 5
autoplot(x)
```

Reading from broad to narrow tells the story of the Big Five: a single general
factor at *k* = 1 splits first to separate Neuroticism, then progressively
differentiates until, at *k* = 5, the canonical five traits appear. Factors carry
stable identifiers `m{k}f{j}` (level *k*, factor *j*) used throughout the object.

Each between-level edge is the factor-score correlation computed through Waller's
(2007) closed-form algebra. The primary-parent edge for every factor, strongest
first, is one line:

```{r}
#| label: tbl-edges
#| tbl-cap: "Primary-parent between-level edges for the Big Five hierarchy, sorted by absolute correlation. Each row gives a factor, its primary parent one level up, and their factor-score correlation."
edges <- tidy(x, what = "edges", primary_only = TRUE, sort = "strength")
knitr::kable(edges[, c("from", "to", "r")], digits = 2)
```

An edge near 1.0 marks a dimension that is stable across that step; smaller values
mark where the structure reorganizes. Factor scores for downstream use are one
call away with `augment(x, data = bfi)`, and a fitted hierarchy can score new
observations with `predict()` without refitting.

<!-- ======================================================================= -->
<!-- APPLIED EXAMPLE — the AMH hierarchy, live from forbes2023 (centerpiece). -->
<!-- ======================================================================= -->

# Applied example: the AMH hierarchy

The package's fidelity to the published method is not merely asserted; it is
tested. We reproduce @forbes2023's applied example, a hierarchical analysis of
155 mental-health variables from the Assessing Mental Health (AMH) study. The
correlation matrix ships with the package as `forbes2023` (licensed CC-BY 4.0 by
its author). Following Forbes, we extract from one to ten factors with all-pairs
correlations:

```{r}
#| label: amh-fit
data(forbes2023)
dim(forbes2023)
amh <- ackwards(forbes2023, k_max = 10, pairs = "all")
```

The full ten-level hierarchy (@fig-amh) is far richer than the illustrative
example, with broad internalizing and externalizing dimensions at the top
progressively differentiating into specific syndromes.

```{r}
#| label: fig-amh
#| fig-cap: "The Assessing Mental Health (AMH) bass-ackwards hierarchy, 155 variables extracted from one to ten factors with all-pairs between-level correlations, reproducing @forbes2023. Only edges at or above the display threshold are drawn."
#| fig-width: 9
#| fig-height: 7
autoplot(amh)
```

Piping the result through `prune("redundant")` applies Forbes's redundancy
diagnostic, flagging factors that persist near-unchanged across levels rather
than differentiating:

```{r}
#| label: amh-prune
amh_pruned <- prune(amh, "redundant")
nodes <- tidy(amh_pruned, what = "nodes")
c(total_factors = nrow(nodes), flagged_redundant = sum(nodes$pruned))
```

**Fidelity.** This is the same computation exercised by the package's test suite,
which verifies it against Forbes's own reference implementation (OSF `pcwm8`).
Across all 45 between-level pairs, **ackwards**'s edge correlations reproduce the
reference values to within `r sprintf("%.1e", 1.3e-14)` in absolute magnitude,
and the redundancy chase reproduces her published result for all 54 components
(the direct-correlation criterion **ackwards** adopts as its default). The
applied example a reader renders here is thus numerically identical, to machine
precision, to the analysis reported by @forbes2023.

<!-- ======================================================================= -->
<!-- DISCUSSION — author-owned stub. Draft the closing argument here.         -->
<!-- ======================================================================= -->

# Discussion

> **\[AUTHOR TO DRAFT — structured stub]**
> Suggested beats:
>
> - What the package enables that bespoke scripts did not: engine breadth, a
>   single verified edge path, exact reproduction of the published method, tested
>   and documented software.
> - Interpretive cautions the package deliberately enforces: descriptive not
>   confirmatory hierarchy; reporting fit values rather than pass/fail verdicts;
>   redundancy/artifact flags as guides, not deletions; selection bias in reading
>   the single strongest edge.
> - Scope and limitations: orthogonal rotation only; EAP scoring and higher-order
>   SEM out of scope; convergence behavior at deep levels.
> - Future directions and an availability statement (CRAN, GitHub, documentation
>   site, open-source license).

# Reproducibility

All numeric results, tables, and figures in this manuscript are computed at
render time from the **ackwards** package [@ackwards] running in R [@rcore2026];
no values are hardcoded. See `manuscript/README.md` in the package repository for
the render command, required toolchain, and the package version used.

# Use of generative AI

The **ackwards** package and this manuscript were developed with the assistance
of a large language model (Anthropic's Claude, using its Opus and Sonnet models)
through the Claude Code interface. Rather than a single prompt, the assistant was
used iteratively across many sessions for software implementation, test
construction, documentation, and editing of author-written prose. Because this
mode of use produces no reproducible prompt, we disclose its scope rather than a
transcript: every AI-assisted contribution was reviewed, verified, and edited by
the author, who takes full responsibility for the final code, analyses, and text.
Consistent with journal and publisher policy, no AI system is credited as an
author.

Crucially, AI assistance does not stand in for the manuscript's verification
guarantees. Every numeric result reported here is checked against at least two
independent oracle types — published or closed-form values, an independent
software implementation, or seeded simulation — and the central applied example
reproduces the reference implementation of @forbes2023 to machine precision. The
complete development record, including design decisions, milestone histories, and
the catalogue of verification oracles, is version-controlled in the package
repository and can be inspected independently of the assistant used to produce
it.

# References

::: {#refs}
:::
