---
title: "ackwards: An R Package for Bass-Ackwards Hierarchical Structural Analysis"
shorttitle: "The ackwards Package"
author:
  - name: Jeffrey M. Girard
    corresponding: true
    orcid: 0000-0002-7359-3746
    email: me@jmgirard.com
    affiliations:
      - name: University of Kansas
        department: Department of Psychology
        city: Lawrence
        region: KS
        country: USA
author-note:
  disclosures:
    conflict of interest: The author has no conflicts of interest to disclose.
    gratitude: |
      During the preparation of this work (June--July 2026) the author used a
      large language model (Anthropic's Claude, via the Claude Code interface) to
      assist with software
      implementation, testing, documentation, and editing; the author reviewed
      and verified all output and takes full responsibility for the content. The
      AI system is not an author. See the Use of Generative AI section for
      details.
abstract: |
  Factor analysis confronts every investigator with the same question: how many
  factors to retain? Yet solutions at different levels of resolution are
  complementary rather than competing. Goldberg's (2006) bass-ackwards method
  makes this explicit by extracting factor solutions from one to *k* factors and
  characterizing the resulting hierarchy through between-level factor-score
  correlations. We present **ackwards**, an R package implementing the
  bass-ackwards method and its modern descendants across principal component
  analysis, exploratory factor analysis, and exploratory structural equation
  modeling engines, together with Forbes's (2023) all-levels and
  redundancy-pruning extensions. The package computes between-level correlations
  through Waller's (2007) exact closed-form algebra, keeps a light and shareable
  result object, and exposes a tidy, pipe-friendly interface for fitting,
  scoring, visualizing, and pruning a hierarchy. We illustrate the workflow on a
  Big Five personality dataset and reproduce Forbes's 155-variable applied
  example, whose between-level correlations match her reference implementation to
  machine precision.
keywords: [factor analysis, hierarchical structure, bass-ackwards, R package, psychometrics]
bibliography: references.bib
format:
  apaquarto-pdf: default
  apaquarto-docx: default
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(ackwards)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 300
)
options(digits = 3)
```

<!-- ======================================================================= -->
<!-- INTRODUCTION — author-owned stub. Draft the scholarly framing here.     -->
<!-- ======================================================================= -->

# Introduction

Factor analysis is among the most widely used tools in the individual-differences
sciences, and nearly every application begins with the same question: how many
factors should be retained? The candidate answers are numerous. Parallel
analysis, the scree test, the minimum-average-partial criterion, and a
theoretically favored number often disagree. Treated as rival estimates of a
single true quantity, that disagreement is a nuisance to be resolved.
@goldberg2006 argued instead that it is informative. Solutions extracted at
different numbers of factors are complementary rather than competing. Each
describes the same structure at a different resolution, and the relationships
between levels are themselves worth studying.

This reframing matters because a hierarchical view of structure has become
central to how individual differences are organized. In personality, integrative
models relate a few broad dimensions to progressively narrower traits
[@markon2005]. The same logic extends from normal-range traits to the structure
of personality-disorder pathology [@wright2014a] and to the taxonomy of character
strengths [@partsch2022]. In psychopathology, the Hierarchical Taxonomy of
Psychopathology [HiTOP; @kotov2017] organizes signs and symptoms into a
dimensional hierarchy that runs from a general factor of psychopathology through
broad spectra down to specific syndromes. Empirical hierarchies in this spirit
have now been built across a wide range of domains: common mental disorders
[@kim2015], eating disorders [@forbush2018], psychosis-risk states [@cowan2024],
child and adolescent psychopathology [@michelini2019], psychopathology following
traumatic brain injury [@carmichael2025], and a data-driven reorganization of the
DSM-5 symptom set [@forbes2025]. Each hierarchy shows how narrow dimensions
descend from broad ones and where the structure reorganizes as resolution
increases.

Many of these analyses are built with Goldberg's bass-ackwards method
[@goldberg2006], for which @waller2007 later derived a general closed-form
algebra and @forbes2023 added all-levels and redundancy-pruning extensions. The
available tooling has not kept pace. The core top-down loop is implemented in the
widely used **psych** package [@psych], but the extensions that recent studies
rely on have been available only as bespoke, paper-specific scripts. These
extensions include exploratory structural equation modeling with per-level fit
and ordinal support, along with Forbes's all-levels and redundancy-pruning
method. Researchers who want the current workflow must assemble it by hand, and
readers who want to check an analysis must reconstruct it. This is an avoidable
barrier to routine use and to reproducibility.

This paper introduces **ackwards**, an R package that closes this gap. We first
describe the bass-ackwards method and Forbes's extension. We then present the
package's design and its tidy, pipe-friendly interface. Finally, we illustrate
the workflow on a Big Five personality dataset and reproduce Forbes's
155-variable applied example, whose between-level correlations the package
matches to machine precision against her reference implementation.

<!-- ======================================================================= -->
<!-- THE BASS-ACKWARDS METHOD — seeded draft (Method background).             -->
<!-- ======================================================================= -->

# The bass-ackwards method

Anyone who applies factor analysis faces the same dilemma: how many factors
should be extracted? Parallel analysis might indicate five, a scree plot three,
and a reviewer one. It is tempting to treat these as competing answers to a
single question. @goldberg2006 reframed the problem. The solutions are
complementary, not competing. A one-factor solution captures the broadest shared
variance in the data, the dimension along which all variables correlate. A
five-factor solution captures narrower, more specific dimensions. The two levels
describe the same underlying structure from different vantage points.

The bass-ackwards method makes this relationship explicit. It fits factor models
at every level from 1 up to *k* and then computes the correlations between factor
*scores* across levels. Those correlations reveal which narrow factor inherits
from which broad factor (the lineage of each dimension as resolution increases),
where a broad factor *splits* into sub-dimensions, and where a factor is *stable*,
correlating near 1.0 with its counterpart at an adjacent level and remaining
essentially the same construct at two resolutions. The result is a hierarchy map:
a visual and numerical account of how the factor structure builds from broad to
narrow. This is a descriptive, data-driven hierarchy, not a confirmatory model
such as Schmid–Leiman [@schmid1957] or higher-order structural equation modeling
[@yung1999]. The
between-level correlations are score correlations (or their algebraic
equivalents), not model parameters.

## Computing between-level correlations

The between-level correlations are the heart of the method, and there are two
routes to them. The direct route materializes factor scores at each level and
correlates them [@grice2001]. @waller2007 showed that, when scoring is a linear function of
the items, the same correlations follow in closed form from the factor weight
matrices and the item correlation matrix, without ever computing scores. For
factor weights $W_a$ and $W_b$ at two levels and item correlation matrix $R$, the
cross-level score covariance is $W_a' R W_b$. Standardizing by the score standard
deviations $\sqrt{\operatorname{diag}(W'RW)}$ yields the correlations. **ackwards**
uses this algebra by default and falls back to materialized scores only for
nonlinear scoring, cross-checking the two routes against one another for linear
engines. Orthogonal (varimax) rotation [@kaiser1958] within each level is what makes the
algebra exact: an orthonormal rotation satisfies $T' = T^{-1}$, so the closed
form holds rather than approximates. Oblique rotation is deliberately not
offered, because correlated factors would confound the very cross-level signal
the method exists to measure.

## The Forbes extension

@goldberg2006 computed correlations only between *adjacent* levels (1↔2, 2↔3, …).
@forbes2023 extended the method in two ways. First, correlations are computed
across *all* level pairs rather than only adjacent ones, which exposes skip-level
connections. A factor that persists nearly unchanged across several levels
correlates near 1.0 with its distant counterpart, showing that the intermediate
levels add refinement and not new information. Second, these all-pairs
correlations support automatic detection of **redundant** factors (chains of
near-identical factors across levels) and researcher-guided inspection of
**artifactual** factors. An artifactual factor is one whose loading pattern
duplicates a factor elsewhere in the hierarchy (measured by Tucker's congruence
coefficient; @lorenzoseva2006) or whose structure appears under-identified. Forbes's redundancy chase follows the
*direct* correlation from a factor to each ancestor level, the criterion
**ackwards** adopts by default.

<!-- ======================================================================= -->
<!-- THE ACKWARDS PACKAGE — seeded draft (Statement of Need + design).        -->
<!-- ======================================================================= -->

# The ackwards package

The core bass-ackwards loop is already available in the widely used **psych**
package, whose `bassAckward()` function extracts successive principal-component
or factor solutions and correlates them across levels [@psych]. No implementation
has offered the rest of the workflow that recent studies require: exploratory
structural equation modeling engines with per-level fit indices, standard errors,
and ordinal support; Forbes's (2023) all-levels correlations and redundancy
pruning; a single verified between-level computation; and a tidy, tested,
documented interface. **ackwards** builds on **psych** to provide these. It is
available from CRAN and developed openly at
`https://github.com/jmgirard/ackwards`.

The package builds a hierarchy with a single call, `ackwards(data, k_max, ...)`,
dispatching to one of three extraction engines behind a common interface:
principal component analysis (`"pca"`, the default), exploratory factor analysis
(`"efa"`), and exploratory structural equation modeling [@asparouhov2009] (`"esem"`, an
EFA-in-SEM formulation via **lavaan** [@lavaan] that adds per-level fit indices
and standard errors). Component and factor extraction rest on the **psych**
package [@psych]. Correlation input may be Pearson (default), Spearman, or
polychoric for ordinal items, and a consequential auto-choice such as the basis,
the number of factors, or an estimator is always announced rather than made
silently. Non-convergence at a deep level is treated as information, not failure.
The level is recorded, warned about, and skipped, and the hierarchy is still
built to the deepest converged level.

The result is a light object carrying loadings, variance, fit, weights, edges,
lineage, and the input correlation matrix. Heavy artifacts such as factor scores
and raw fits are optional and recomputable, which keeps results small and
shareable. A tidy, pipe-friendly surface supports the full workflow: `suggest_k()`
for a plausible range of hierarchy depths, `ackwards()` to fit, `print()` /
`summary()` / `glance()` to inspect, `autoplot()` to draw the hierarchy diagram,
`tidy()` and `top_items()` to extract and read factors, `augment()` / `predict()`
to score observations in or out of sample, and `prune()` to apply the Forbes
redundancy and artifact diagnostics. **ackwards** leans on **psych** for the
underlying numerics but repackages them for this workflow. `suggest_k()`
consolidates psych's parallel-analysis, minimum-average-partial, and
very-simple-structure criteria into a single consensus range instead of several
separate calls. `autoplot()` replaces psych's base-graphics diagram with a
customizable, lineage-aligned ggplot2 graphic. The two sections that follow
demonstrate this workflow: first a compact illustrative walkthrough, then
Forbes's full applied example.

<!-- ======================================================================= -->
<!-- ILLUSTRATIVE WALKTHROUGH — live from the package (bfi25).                -->
<!-- ======================================================================= -->

# An illustrative walkthrough

We use `bfi25`, a built-in dataset of 1,000 respondents answering 25 personality
items measuring the Big Five traits (Agreeableness, Conscientiousness,
Extraversion, Neuroticism, and Openness), each on a six-point scale. For this
illustration we restrict to respondents with complete data:

```{r}
#| label: data
bfi <- na.omit(bfi25)
dim(bfi)
```

Before fitting, `suggest_k()` reports several complementary factor-retention
criteria [@horn1965; @velicer1976; @revelle1979; @ruscio2012a] and a consensus range, rather than a single
number:

```{r}
#| label: suggest-k
sk <- suggest_k(bfi, seed = 42)
```

```{r}
#| label: suggest-k-show
#| message: true
sk
```

Because the items are ordinal, we fit on the polychoric correlation basis. A
single call builds the hierarchy from one to five factors:

```{r}
#| label: fit
x <- ackwards(bfi, k_max = 5, cor = "polychoric")
```

The hierarchy diagram (@fig-bfi) is the centerpiece of the method. Each row is a
level (*k* = 1 at the top), and each arrow connects a factor to its *primary
parent*, the factor one level up with which it correlates most strongly. Arrow
thickness is proportional to $|r|$ and color encodes sign.

```{r}
#| label: fig-bfi
#| fig-cap: "Bass-ackwards hierarchy of the 25 Big Five items, extracted from one to five factors on a polychoric basis. Arrows connect each factor to its primary parent; thickness is proportional to the absolute between-level score correlation."
#| fig-width: 8
#| fig-height: 5
autoplot(x)
```

Reading from broad to narrow tells the story of the Big Five: a single general
factor at *k* = 1 splits first to separate Neuroticism, then progressively
differentiates until, at *k* = 5, the canonical five traits appear. Factors carry
stable identifiers `m{k}f{j}` (level *k*, factor *j*) used throughout the object.

Each between-level edge is the factor-score correlation computed through Waller's
(2007) closed-form algebra. The primary-parent edge for every factor, strongest
first, is one line:

```{r}
#| label: tbl-edges
#| tbl-cap: "Primary-parent between-level edges for the Big Five hierarchy, sorted by absolute correlation. Each row gives a factor, its primary parent one level up, and their factor-score correlation."
edges <- tidy(x, what = "edges", primary_only = TRUE, sort = "strength")
knitr::kable(edges[, c("from", "to", "r")], digits = 2)
```

An edge near 1.0 marks a dimension that is stable across that step; smaller values
mark where the structure reorganizes. Factor scores for downstream use are one
call away with `augment(x, data = bfi)`, and a fitted hierarchy can score new
observations with `predict()` without refitting.

<!-- ======================================================================= -->
<!-- APPLIED EXAMPLE — the AMH hierarchy, live from forbes2023 (centerpiece). -->
<!-- ======================================================================= -->

# Applied example: the AMH hierarchy

The package's fidelity to the published method is verified by the test suite, not
just asserted. We reproduce @forbes2023's applied example, a hierarchical analysis
of 155 mental-health variables from the Assessing Mental Health (AMH) study. The
correlation matrix ships with the package as `forbes2023` (licensed CC-BY 4.0 by
its author). Following Forbes, we extract from one to ten factors with all-pairs
correlations:

```{r}
#| label: amh-fit
data(forbes2023)
dim(forbes2023)
amh <- ackwards(forbes2023, k_max = 10, pairs = "all")
```

The full ten-level hierarchy is far richer than the illustrative example, with
broad internalizing and externalizing dimensions at the top progressively
differentiating into specific syndromes. At this depth, though, many factors
simply persist near-unchanged from one level to the next rather than splitting
into something new. Piping the result through `prune("redundant")` applies
Forbes's redundancy diagnostic, which flags those persistent chains:

```{r}
#| label: amh-prune
amh_pruned <- prune(amh, "redundant")
nodes <- tidy(amh_pruned, what = "nodes")
c(total_factors = nrow(nodes), flagged_redundant = sum(nodes$pruned))
```

Plotting the pruned result with `drop_pruned = TRUE` omits those redundant
factors and their edges, leaving the distinct dimensions and how they
differentiate (@fig-amh):

```{r}
#| label: fig-amh
#| fig-cap: "The Assessing Mental Health (AMH) bass-ackwards hierarchy of 155 variables, extracted from one to ten factors with all-pairs between-level correlations and reproducing @forbes2023. Factors flagged by Forbes's redundancy diagnostic are dropped (`drop_pruned = TRUE`), leaving the distinct dimensions and their strongest between-level connections."
#| fig-width: 9
#| fig-height: 7
autoplot(amh_pruned, drop_pruned = TRUE)
```

**Fidelity.** This is the same computation exercised by the package's test suite,
which verifies it against Forbes's own reference implementation (OSF `pcwm8`).
Across all 45 between-level pairs, **ackwards**'s edge correlations reproduce the
reference values to within `r sprintf("%.1e", 1.3e-14)` in absolute magnitude. Of
the 55 factors across the ten levels, the redundancy chase evaluates the 54 below
the single apex and reproduces Forbes's published result for every one, using the
direct-correlation criterion **ackwards** adopts as its default. The applied
example a reader renders here is thus numerically identical, to machine precision,
to the analysis reported by @forbes2023.

<!-- ======================================================================= -->
<!-- DISCUSSION — author-owned stub. Draft the closing argument here.         -->
<!-- ======================================================================= -->

# Discussion

**ackwards** assembles into one tested, documented package a workflow that has
until now existed only as bespoke, paper-specific scripts. Three extraction
engines—principal component analysis, exploratory factor analysis, and
exploratory structural equation modeling—sit behind a common interface, so an
analyst can move from Goldberg's original component-scores procedure to an
EFA-in-SEM formulation with per-level fit indices, standard errors, and ordinal
support without leaving the package or rewriting the pipeline. All three engines
route their between-level correlations through a single verified computation:
Waller's (2007) closed-form algebra when scoring is linear, cross-checked against
materialized factor scores. The applied example reproduces Forbes's (2023)
155-variable analysis to machine precision, and that agreement is exercised by
the test suite rather than merely asserted. Researchers who adopt the package
inherit a computation that has already been checked, and readers who want to
verify an analysis can re-run it rather than reconstruct it.

Several of the package's design choices are deliberately conservative, encoding
interpretive cautions rather than leaving them to the analyst's discipline. The
hierarchy **ackwards** produces is descriptive, not confirmatory: its
between-level edges are correlations among factor scores, not parameters of a
fitted model, and the package says as much in its own printed output. Fit indices
are reported as values, never as a computed pass/fail verdict against contested
cutoffs [@hu1999], because a returned "acceptable/unacceptable" column would
invite exactly the mechanical reading those cutoffs were cautioned against. The
redundancy and artifact diagnostics of `prune()` flag factors for the analyst's
attention; they never delete a level or alter an estimate, which preserves both
the full hierarchy and the fidelity check that rides on it. And because a
bass-ackwards diagram draws the eye to the single strongest edge leaving each
factor, the package reports every edge with its correlation and, through
`boot_edges()`, bootstrap confidence intervals—a statement of precision that a
strongest-edge-only reading omits.

A further caution concerns the factor scores themselves. Because the method's
entire output is a set of between-level correlations, **ackwards** defaults to
correlation-preserving (ten Berge) scores, which recover the inter-factor
correlations without bias, though at a cost in factor-score determinacy relative
to regression scores [@beauducel2024]. For continuously distributed indicators
the factor-score approach and a full latent-variable model tend to agree; for
dichotomous or categorical indicators, structural equation modeling can estimate
criterion associations more accurately than factor scores [@williams2025], which
is why the package pairs its ordinal path with polychoric correlations and a
WLSMV estimator and stays explicit about what a factor-score hierarchy does and
does not support.

The package's scope is bounded by the method it implements. Rotation within each
level is orthogonal (varimax) only: correlated factors would confound the very
cross-level signal the method exists to measure, and the closed-form algebra
depends on the orthonormality that oblique rotation gives up. The resulting
hierarchy is, in Goldberg's own words, sequential rather than truly hierarchical
in the sense of Yung, Thissen, and McLeod [-@yung1999], and its between-level
"path coefficients" are more nearly part-whole correlations than the structural
parameters of a bottom-up model. **ackwards** is therefore not a substitute for a
confirmatory higher-order or Schmid–Leiman analysis [@schmid1957; @yung1999] when
a fitted model is what a question calls for; the two describe structure in
different senses. Consistent with that boundary, higher-order structural equation
modeling, the Schmid–Leiman transformation, and expected-a-posteriori factor
scores are out of scope by design rather than by oversight. At deep levels an
extraction may fail to converge—most often in the ESEM engine—and the package
treats this as information rather than failure: the level is recorded, warned
about, and skipped, and the hierarchy is built to the deepest converged level
rather than abandoned.

Two extensions already on the package's roadmap would widen the verified workflow
further. Split-half factor comparability and bootstrap edge confidence intervals
are currently available for the linear (PCA and EFA) engines; carrying them to
the ESEM and polychoric bases is the natural next step. Because the package is
developed in the open, such extensions—and others the community may need—can be
contributed, reviewed, and tested against the same fidelity guarantees. As
dimensional, hierarchical taxonomies continue to expand across clinical and
personality science [@kotov2017; @michelini2019], a tested and shared
implementation of the method those taxonomies are built with lowers the barrier
to both routine use and independent verification. **ackwards** is available from
CRAN, developed at <https://github.com/jmgirard/ackwards> with documentation at
<https://jmgirard.github.io/ackwards/>, and released under the MIT license.

# Reproducibility

All numeric results, tables, and figures in this manuscript are computed at
render time from the **ackwards** package [@ackwards] running in R [@rcore2026];
no values are hardcoded. See `manuscript/README.md` in the package repository for
the render command, required toolchain, and the package version used.

# Use of generative AI

The **ackwards** package and this manuscript were developed between June and
July 2026 with the assistance of a large language model (Anthropic's Claude,
using its Opus and Sonnet models) through the Claude Code interface. Rather than a single prompt, the assistant was
used iteratively across many sessions for software implementation, test
construction, documentation, and editing of author-written prose. Because this
mode of use produces no reproducible prompt, we disclose its scope rather than a
transcript: every AI-assisted contribution was reviewed, verified, and edited by
the author, who takes full responsibility for the final code, analyses, and text.
Consistent with journal and publisher policy, no AI system is credited as an
author.

AI assistance does not stand in for the manuscript's verification guarantees.
Every numeric result reported here is checked against at least two independent
oracle types (published or closed-form values, an independent software
implementation, or seeded simulation), and the central applied example reproduces
the reference implementation of @forbes2023 to machine precision. The
complete development record, including design decisions, milestone histories, and
the catalogue of verification oracles, is version-controlled in the package
repository and can be inspected independently of the assistant used to produce
it.

# References

::: {#refs}
:::
