---
title: "Replicability-Gated Hierarchies: A Recommended Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Replicability-Gated Hierarchies: A Recommended Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

The other vignettes in this package document its individual verbs. This one
takes a position: it lays out the workflow we recommend for answering the
questions bass-ackwards analysis exists to answer, and explains why each step
is there. We refer to it as the **replicability-gated** (or Girard) workflow,
because its central move is to let *replication*, not fit, decide how deep the
hierarchy you interpret goes.

## The problem: depth is where bass-ackwards analyses go wrong

A bass-ackwards analysis asks two questions: *how does the structure of this
domain organize hierarchically*, and *how deep is that hierarchy meaningful?*
The first question is what the method computes. The second is where published
applications most often stumble, and the failure mode is consistent:
**overextraction**, followed by substantive interpretation of deep-level
factors that would not re-emerge in a new sample. Forbes (2023) documents this
directly for the bass-ackwards context — non-replicable structure concentrates
in the deeper levels of an overextracted hierarchy.

The standard tooling half-addresses this. Retention criteria like parallel
analysis (`suggest_k()`) estimate a *plausible range* for k from the
eigenstructure, and they are the right first screen. But
they do not measure the thing the caution is actually about: whether the
specific factors in *your* solution would show up again in another sample from
the same population. A level can sit comfortably inside the plausible range
and still contain a factor that is pure sample idiosyncrasy.

There was, historically, a direct instrument for exactly this. Everett (1983)
proposed **factor comparability coefficients**: split the sample in half, fit
the solution in each half independently, apply *both* halves' scoring weights
to the full sample, and correlate the matched factor scores. A factor that is
real re-emerges in both halves and its two score estimates correlate near 1; a
factor that is noise does not. Goldberg's lab used this as a routine gate —
Goldberg (1990) retained factor solutions only when they held up across
repeated random split-halves — and the practice runs through the research
program that produced the bass-ackwards method itself (Goldberg, 2006). The
modern bass-ackwards literature largely dropped it.

`comparability()` restores that gate as a first-class verb, extended to the
bass-ackwards setting: coefficients for every factor at every level, computed
through the same score-correlation algebra as the hierarchy's own edges.

## Three verbs, three different questions

The workflow below leans on the fact that this package now separates three
questions that are easy to conflate:

| Question | Verb | What it measures |
|---|---|---|
| What depth range is *plausible*? | `suggest_k()` | Eigenstructure retention criteria (consensus range) |
| Which factors *replicate*? | `comparability()` | Split-half score comparability per factor per level |
| Which factors *differentiate*? | `prune()` | Forbes (2023) redundancy: factors that persist without adding resolution |

These are genuinely different. A factor can replicate perfectly and still be
redundant (the same construct restated at every level); a level can sit inside
the plausible range and still fail to replicate. You need all three answers,
and no one of them substitutes for another.

## Setup

We use the `bfi25` Big Five data (25 items, n = 875 after removing incomplete
rows), analyzed with the default PCA engine on the Pearson basis throughout.
That basis is deliberate: `comparability()` and `boot_edges()` screen on
Pearson/Spearman by design — polychoric estimation in every split-half or
bootstrap resample is slow and unstable — so the ordinal-detection warning
`bfi25` triggers is *expected* here and suppressed below for readability. You
screen structure and depth on Pearson, then fit the model you actually *report*
with `cor = "polychoric"`.

```{r data}
library(ackwards)
bfi <- na.omit(bfi25)
```

## Step 1: Screen the plausible depth range with `suggest_k()`

```{r suggest-k, warning = FALSE}
sk <- suggest_k(bfi, seed = 1)
print(sk)
```

```{r sk-inline, include = FALSE}
sk_ks <- na.omit(c(
  sk$k_parallel_pc, sk$k_parallel_fa, sk$k_map, sk$k_vss1, sk$k_vss2,
  if (isTRUE(sk$cd_available)) sk$k_cd
))
sk_lo <- min(sk_ks)
sk_hi <- max(sk_ks)
```

The criteria disagree — they always do — but they bracket a range (here
k = `r sk_lo`–`r sk_hi`). Treat the top of that range as a ceiling worth
*probing*, not a depth worth *asserting*: parallel analysis on the PC basis in
particular tends to overextract, and `vignette("ackwards-suggest-k")` explains
each criterion's bias direction in detail. Deliberately looking one or two
levels past the consensus is a normal part of the method — the point of the
next step is that you can now do so safely.

## Step 2: Gate the depth on replicability with `comparability()`

This is the step the workflow is named for. We evaluate every level up to one
past the ceiling, across ten random split-halves:

```{r comparability, warning = FALSE}
cmp <- comparability(bfi, k_max = sk_hi + 1, n_splits = 10, seed = 2026)
print(cmp)
```

```{r cmp-inline, include = FALSE}
lv_min <- tapply(cmp$summary$r_median, cmp$summary$level, min)
gate <- max(as.integer(names(lv_min))[cumsum(lv_min < 0.9) == 0])
worst_top <- cmp$summary[cmp$summary$level == cmp$k_max, ]
worst_top <- worst_top[which.min(worst_top$r_median), ]
fmt <- function(v) sub("^(-?)0\\.", "\\1.", sprintf("%.2f", v))
```

Read the per-level minima, not the medians: a level is only as interpretable
as its least replicable factor. Here every factor through k = `r gate` clears
the conventional floor (the weakest median comparability at any of those
levels is `r fmt(min(lv_min[seq_len(gate)]))`), and the structure degrades
past it — at k = `r cmp$k_max`, `r worst_top$factor` reaches a median of only
`r fmt(worst_top$r_median)` and dips to `r fmt(worst_top$r_min)` on its worst
split. Those deep factors are not stable dimensions of the data; they are
different factors in every half-sample that happen to occupy the same slot.

```{r comparability-plot, fig.height = 6}
autoplot(cmp)
```

Two honesty notes. First, the conventional benchmarks marked on the plot
(.90 and .95, following Everett's and Goldberg's practice) are conventions,
not tests — `comparability()` reports every coefficient and flags nothing, so
borderline cases stay visible and the judgment stays yours. Second,
comparability is *internal* replication: it asks whether the structure
re-emerges in halves of the *same* sample. That is the right gate for "is this
factor real here," and a necessary condition for — but not a demonstration
of — replication in a new population.

The result is a **hierarchy floor**: the deepest level at which every factor
replicates. For these data that is k = `r gate` — the Big Five, as it should
be.

## Step 3: Fit the hierarchy to the gated depth

```{r fit}
x <- ackwards(bfi, k_max = 5)
print(x)
```

```{r hierarchy-plot}
autoplot(x)
```

If you prefer to *display* the fragmentation beyond the floor (it can be
substantively interesting to show a factor dissolving), fit one level deeper
and say explicitly in your report which levels passed the replicability gate.
What the gate rules out is interpreting the deeper factors as constructs.

## Step 4: Find what perpetuates without differentiating — `prune()`

Replicability is necessary but not sufficient: a factor can re-emerge in every
half-sample and still add nothing, because it is the same construct restated
level after level. That is Forbes's (2023) redundancy question, and it is
`prune()`'s job, not `comparability()`'s:

```{r prune}
pr <- prune(x, "redundant")
pr$prune$chains
```

Here the chains tell us that several of the Big Five arrive early and simply
persist — for example one factor emerges at k = 2 and travels essentially
unchanged (r ≥ .97 at every link) down to k = 5. Every node on that chain
*replicates* beautifully; the chain flags that the intermediate appearances
add no resolution. The two verbs answer different questions, and reporting
both gives the honest picture: which levels are real, and which levels are
new. See `vignette("ackwards-forbes")` for chain mechanics, thresholds, and
the retention rule.

### How sure are we about these edges? `boot_edges()`

`prune()`'s redundancy rule and the Forbes practice of reading off the
strongest skip-level correlation both consume a point estimate at a hard
threshold, so the natural follow-up is how precisely each edge is estimated.
`boot_edges()` attaches a nonparametric bootstrap standard error and percentile
confidence interval to every edge. Each replicate resamples respondents,
recomputes the correlations, refits the whole hierarchy, and — importantly —
re-anchors each replicate's factors to the full-sample solution (the same
matching and sign-orientation `comparability()` uses), so factor
label-switching across replicates does not contaminate the intervals. It runs
on the PCA and EFA engines with a Pearson or Spearman basis — a polychoric
matrix is too slow and unstable to re-estimate in every replicate (the same
scope as `comparability()`), so screen edge stability on the Pearson basis even
when the final model is polychoric.

```{r boot-edges}
x_boot <- ackwards(bfi, k_max = 5, pairs = "all") |>
  boot_edges(bfi, n_boot = 200, seed = 1)

boot_tbl <- tidy(x_boot, what = "edges", sort = "strength")
skip_boot <- boot_tbl[
  abs(boot_tbl$level_to - boot_tbl$level_from) > 1 & abs(boot_tbl$r) >= 0.5,
  c("from", "to", "r", "lo", "hi")
]
head(skip_boot, 6)
```

The `lo`/`hi` columns are the 95% percentile interval. A skip-level edge whose
interval sits comfortably above `prune()`'s `redundancy_r` threshold (0.9 by
default) is a defensible "same construct" claim; one whose interval straddles
the threshold should not be treated as decisively redundant. The intervals are
per-edge error bars, not a familywise guarantee: they do not correct for having
*searched* the full edge table for the largest value, so a strongest-edge claim
should still be pre-specified rather than selected after the fact.

## Step 5: Interpret with the built-in guardrails

```{r interpret}
top_items(x, level = 5, cut = 0.5)
```

Three interpretation habits keep the analysis honest (all covered in depth in
`vignette("ackwards-interpret")`):

* **Negative loadings do not mean "low"** — sign is anchored to the primary
  parent, and a negative secondary edge is information, not error.
* **Name factors within levels, borrowing names down the lineage** rather than
  treating each level as a fresh exploratory result.
* **Say what the hierarchy is.** A bass-ackwards result is a series of linked
  solutions whose edges are score correlations — it is descriptive, not a
  fitted hierarchical model (no Schmid–Leiman, no higher-order SEM), and
  should be reported as such.

## Step 6: Validate downstream, out of sample

The final step closes the loop the same way it opened: with data the model
has not seen. Scores for new observations use the fit-time metric, so
training and test scores are directly comparable:

```{r train-test}
train <- bfi[1:600, ]
test <- bfi[601:875, ]

x_train <- ackwards(train, k_max = 5)
test_scores <- predict(x_train, newdata = test)
round(head(test_scores[, grep("^\\.m5", names(test_scores))], 3), 2)
```

If a downstream claim (a correlation with an outcome, a group difference)
holds for training-sample scores but not test-sample scores, the hierarchy was
fine and the claim was overfit — a different failure than anything in Steps
1–5, and one only held-out data can catch. The train/test mechanics are
covered in `vignette("ackwards-intro")`.

## Common mistakes this workflow prevents

**Interpreting non-replicable deep factors.** The classic failure. Without a
replicability instrument the deep levels are just *there*, printed with the
same authority as the robust ones. Step 2 makes their status measurable and
visible.

**Trusting a single split.** One split-half can flatter or slander a factor by
luck of the draw. `comparability()` defaults to ten random splits and reports
the spread; the minima across splits are as informative as the medians.

**Cherry-picking the strongest edge.** With `pairs = "all"`, the number of
between-level correlations grows fast, and reporting the most dramatic one
capitalizes on chance. Report full edge tables (or the complete diagram), and
treat any "strongest link" claim as descriptive rather than inferential.
`boot_edges()` (Step 4 above) attaches a bootstrap confidence interval to every
edge, which makes each estimate's precision visible — useful for judging whether
an edge clears `prune()`'s redundancy threshold — but per-edge intervals do not
undo the selection bias of scanning many edges for the largest.

**Mistaking persistence for structure.** A chain of near-1.0 correlations down
the hierarchy looks impressive but means the levels are restating one
construct. That is a `prune()` finding, and it should shorten your reported k
range, not lengthen your list of constructs.

**Mistaking replicability for validity.** A comparability of .99 means the
factor is a stable feature of these items in this population — not that it is
a good construct, and not that it generalizes beyond the population sampled.
Cross-population replication and external validation remain separate,
subsequent steps.

## Summary: the workflow in six calls

1. **`suggest_k(data)`** — screen the plausible depth range; treat its top as
   a ceiling to probe.
2. **`comparability(data, k_max = ceiling + 1)`** — find the hierarchy floor:
   the deepest level at which every factor replicates across split-halves.
3. **`ackwards(data, k_max = floor)`** — fit the hierarchy to the gated depth.
4. **`prune(x, "redundant")`** — identify factors that perpetuate without
   differentiating; report which levels add resolution.
5. **`top_items(x)` + `autoplot(x)`** — interpret with the sign, naming, and
   descriptive-hierarchy guardrails.
6. **`predict(x_train, newdata = test)`** — validate downstream claims on
   held-out data in the fit-time metric.

Steps 1, 3, 4, and 5 are the established method. Step 2 restores the gate the
method's own inventor applied to it, and Step 6 extends the same logic — *show
me it holds in data you did not fit* — to whatever you do with the scores.

## References

Everett, J. E. (1983). Factor comparability as a means of determining the
number of factors and their rotation. *Multivariate Behavioral Research*,
*18*(2), 197–218. https://doi.org/10.1207/s15327906mbr1802_5

Forbes, M. K. (2023). Improving hierarchical models of individual differences:
An extension of Goldberg's bass-ackward method. *Psychological Methods*.
https://doi.org/10.1037/met0000546

Goldberg, L. R. (1990). An alternative "description of personality": The
Big-Five factor structure. *Journal of Personality and Social Psychology*,
*59*(6), 1216–1229. https://doi.org/10.1037/0022-3514.59.6.1216

Goldberg, L. R. (2006). Doing it all bass-ackwards: The development of
hierarchical factor structures from the top down. *Journal of Research in
Personality*, *40*(4), 347–358. https://doi.org/10.1016/j.jrp.2006.01.001

Lorenzo-Seva, U., & ten Berge, J. M. F. (2006). Tucker's congruence coefficient
as a meaningful index of factor similarity. *Methodology*, *2*(2), 57–64.
https://doi.org/10.1027/1614-2241.2.2.57
