---
title: "Introduction to Bass-Ackwards Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Bass-Ackwards Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

## The problem: which factor solution is right?

Anyone who has applied factor analysis faces the same dilemma: how many factors
should you extract? Parallel analysis might say 5; the scree plot might suggest
3; a reviewer might insist on 1. Each solution seems to describe the data
differently, and it is tempting to treat them as competing answers to the same
question.

Goldberg (2006) reframed this problem: the different solutions are not competing
— they are complementary. A 1-factor solution captures the broadest shared
variance in the data, the dimension along which all variables correlate. A
5-factor solution captures narrower, more specific dimensions. These two levels
of resolution describe the same underlying structure from different vantage
points, just as a satellite image and a street map describe the same city.

The **bass-ackwards method** makes this relationship explicit. It fits factor
models at every level from 1 up to k, and then computes the correlations between
factor *scores* across adjacent levels. Those correlations show you:

- **Which narrow factor inherits from which broad factor** — the lineage of
  each dimension as you move from coarse to fine resolution.
- **Where a factor splits** — a single broad factor that correlates strongly
  with two narrower factors is fragmenting into sub-dimensions.
- **Where a factor is stable** — a factor that correlates ~1.0 with its
  counterpart at the adjacent level is essentially the same construct appearing
  at two resolutions.

The result is a **hierarchy map** — a visual and numerical account of how
the factor structure of your data builds from broad to narrow.

> *Note: This is a descriptive, data-driven hierarchy, not a confirmatory
> hierarchical model like Schmid–Leiman or higher-order SEM. The between-level
> correlations are score correlations (or their algebraic equivalents), not
> model parameters. See `vignette("engines")` for when to use EFA or ESEM
> instead of PCA.*

## Data

We use the 25-item Big Five Inventory from the `psych` package (`psych::bfi`).
The items measure five personality traits — Agreeableness (A1–A5),
Conscientiousness (C1–C5), Extraversion (E1–E5), Neuroticism (N1–N5), and
Openness (O1–O5) — each on a 6-point Likert scale, in 2,800 participants.

```{r data}
library(ackwards)

bfi <- na.omit(psych::bfi[, 1:25])
dim(bfi)
```

We drop 364 cases with any missing item, leaving n = 2,436. In a real analysis
you might use pairwise deletion or multiple imputation; `na.omit()` is
sufficient for this illustration.

## Step 1: How many factors? `suggest_k()`

Before fitting the hierarchy, it helps to have a sense of the plausible range
of k. `suggest_k()` runs two established selection criteria:

- **Parallel analysis** (Horn, 1965): compares observed eigenvalues against
  those from random data of the same size and suggests retaining factors whose
  observed eigenvalues exceed the parallel-analysis threshold.
- **Velicer's MAP criterion** (Velicer, 1976): examines the average
  off-diagonal partial correlation after successive factors are extracted;
  suggests the k at which this average is minimized.

```{r suggest_k}
sk <- suggest_k(bfi)
sk
```

> **What is that warning?** `suggest_k()` detected that the BFI items look
> ordinal (integer-valued, ≤ 7 distinct values). By default it still uses
> Pearson correlations, which can underestimate factor loadings for ordinal
> items. When you move to `ackwards()`, you can address this with
> `cor = "polychoric"` — we will do that next.

Both criteria agree: k = 5. This matches the instrument design (Big Five) and
gives us a clear target. The note at the bottom reminds us that k in
`ackwards()` is an *upper bound* — the hierarchy covers every level from 1 to
k, so setting k one or two levels higher than the consensus is often
informative.

## Step 2: Fit the hierarchy `ackwards()`

Now fit the bass-ackwards hierarchy. The most important arguments are:

| Argument | What it controls | Default |
|----------|-----------------|---------|
| `k` | Maximum depth of the hierarchy | *(required)* |
| `method` | Extraction engine: `"pca"`, `"efa"`, or `"esem"` | `"pca"` |
| `cor` | Correlation type: `"pearson"`, `"spearman"`, `"polychoric"` | `"pearson"` |
| `rotation` | Rotation criterion (orthogonal CF ≈ varimax) | `"cfT"` |

Because BFI items are ordinal, we use `cor = "polychoric"`. This computes
polychoric correlations between items before factor extraction, giving a more
accurate representation of the latent structure.

```{r fit, warning = FALSE}
x <- ackwards(bfi, k = 5, cor = "polychoric")
```

No warning this time — specifying `cor = "polychoric"` tells `ackwards()` that
you have already thought about the measurement scale.

## Step 3: Summarize the result

### High-level summary

`print()` gives a quick overview:

```{r print}
x
```

The "Levels" section confirms that all five models converged, and reports the
cumulative variance explained at each level. Notice that the jump from k = 1 to
k = 2 is large (22.9% → 34.7%), while later jumps are smaller — characteristic
of data with a strong general factor and several specific dimensions.

The "Edges" section reports how many of the 40 possible between-level
connections exceed the display threshold of |r| ≥ 0.3. The 40 comes from
summing across all adjacent pairs: 1×2 + 2×3 + 3×4 + 4×5 = 2 + 6 + 12 + 20 =
40.

`glance()` returns the same information as a one-row data frame, convenient for
comparisons across models:

```{r glance}
glance(x)
```

## Step 4: Visualize the hierarchy `autoplot()`

The hierarchy diagram is the centerpiece of the method. Each column represents
one level (k = 1 on the left, k = 5 on the right). Arrows connect each factor
to its **primary parent** — the factor at the level above with which it has the
strongest correlation (|r|). Arrow thickness is proportional to |r|; color shows
direction (blue = positive, orange = negative by default); solid arrows indicate
strong connections (|r| ≥ 0.6), dashed arrows show moderate ones (|r| ≥ 0.3).

```{r autoplot, fig.width = 8, fig.height = 5, warning = FALSE}
autoplot(x)
```

Reading this diagram from left to right tells the story of the Big Five:

- **k = 1**: One broad factor. Given the polychoric basis, this captures shared
  variance across all 25 items.
- **k = 2**: The broad factor splits into two: one that will become
  Extraversion/Agreeableness/Conscientiousness and one that will become
  Neuroticism/Openness.
- **k = 3**: Three factors emerge, with Neuroticism separating out.
- **k = 4**: Agreeableness and Extraversion begin to differentiate.
- **k = 5**: The canonical Big Five appear.

The factors are labeled `m{k}f{j}` (level k, factor j). These stable IDs are
used throughout the object, so `m5f1` at k = 5 refers to the same factor in
the loadings, the edge table, the factor scores, and the diagram.

### Adjusting the diagram

```{r autoplot-custom, fig.width = 8, fig.height = 5, warning = FALSE}
# Show only strong connections; relabel the k = 5 factors
autoplot(x, cut_show = 0.5, cut_strong = 0.7)
```

See `?autoplot.ackwards` for the full argument list.

## Step 5: Dig into the factors `tidy()`

`tidy()` extracts any component of the result as a tidy data frame. The `what`
argument controls what is returned.

### Factor loadings

```{r tidy-loadings}
loadings_df <- tidy(x, what = "loadings")
head(loadings_df)
```

Each row is one item at one level for one factor. You can filter to a specific
level to see what items define each factor. Here are the strongest-loading items
at k = 5:

```{r top-loadings}
lv5 <- loadings_df[loadings_df$level == 5, ]
lv5 <- lv5[order(-abs(lv5$loading)), ]
lv5[!duplicated(lv5$factor), c("factor", "item", "loading")]
```

### Between-level edges

```{r tidy-edges}
edges <- tidy(x, what = "edges")
# Primary-parent edges sorted by strength
primary <- edges[edges$is_primary, c("from", "to", "r")]
primary[order(-abs(primary$r)), ]
```

`is_primary` marks the strongest-connecting edge for each factor (its primary
parent). The r values close to 1.0 indicate factors that are nearly identical
across adjacent levels — a sign of a stable, replicable dimension. Smaller
values indicate where the structure is reorganizing.

### Variance explained

```{r tidy-variance}
tidy(x, what = "variance")
```

Each row is one factor at one level. `variance_pct` is the percentage of total
item variance explained by that factor; `cumulative_pct` accumulates within a
level.

## Step 6: Score observations `augment()`

Factor scores place every observation on each factor at every level. They are
useful for regression, clustering, or any downstream analysis where you want a
continuous summary of a latent dimension.

`augment(x, data = bfi)` computes scores on the fly from the stored weight
matrices and appends them to your data frame. The score columns are named
`.m{k}f{j}` to distinguish them from the original variables.

```{r augment, warning = FALSE}
scored <- augment(x, data = bfi)
dim(scored)         # 25 original items + 15 score columns (1+2+3+4+5)
names(scored)[26:40]
```

Scores are standardized so that the **model-implied** variance is 1 — meaning
the scaling comes from the polychoric correlation matrix, not from the raw data.
In practice the empirical standard deviations will be close to but not exactly 1
when a non-Pearson basis is used; see `?augment.ackwards` for details. For pure
PCA on Pearson correlations the model-implied and empirical variances agree
exactly.

If you need scores repeatedly (e.g., in a cross-validation loop), store them at
fit time with `scores = TRUE` to avoid re-applying the weight matrices on every
call:

```{r scores-stored, warning = FALSE}
x2 <- ackwards(bfi, k = 5, cor = "polychoric", scores = TRUE)
scored2 <- augment(x2)   # uses stored matrices — no recomputation
identical(round(scored2$.m5f1, 8), round(scored$.m5f1, 8))
```

### Using scores for downstream analysis

Because `augment()` returns a plain data frame, the scores slot directly into
any standard R workflow. As a quick illustration, the k = 5 factor scores should
be nearly uncorrelated with each other — a consequence of orthogonal rotation —
while scores across levels should be highly correlated along the primary-parent
lineage:

```{r downstream, warning = FALSE}
# Within-level correlations at k = 5: should be near zero (orthogonal rotation)
k5 <- scored[, grep("^\\.m5", names(scored))]
round(cor(k5), 2)
```

```{r downstream2, warning = FALSE}
# Cross-level: m4f1 is the primary parent of m5f2 and m5f4 (from the edge table)
round(cor(scored[, c(".m4f1", ".m5f1", ".m5f2", ".m5f4")]), 2)
```

The cross-level block confirms lineage: `.m4f1` correlates strongly with
`.m5f2` and `.m5f4` (the k = 5 factors it spawned) and near-zero with `.m5f1`
(which descends from a different k = 4 factor). This is a sanity check you can
run on any result — strong parent–child correlations should appear exactly where
the `tidy(what = "edges")` table says they should.

## Summary

The bass-ackwards workflow in **ackwards** has six steps:

1. **`suggest_k(data)`** — identify a plausible range for the hierarchy depth.
2. **`ackwards(data, k, cor = ...)`** — fit the full hierarchy of factor models.
3. **`print(x)` / `glance(x)`** — check that all levels converged and read the
   top-level summary.
4. **`autoplot(x)`** — visualize the hierarchy as a lineage diagram.
5. **`tidy(x, what = ...)`** — extract loadings, edges, or variance in a form
   ready for tables or further analysis.
6. **`augment(x, data = ...)`** — generate factor scores for downstream use.

## Next steps

| Topic | Vignette |
|-------|---------|
| When PCA is not enough: comparing EFA and ESEM engines | `vignette("ackwards-engines")` |
| Ordinal data: polychoric correlations and WLSMV estimation | `vignette("ackwards-ordinal")` |
| Skip-level connections and pruning with the Forbes extension | `vignette("ackwards-forbes")` |

## References

Goldberg, L. R. (2006). Doing it all bass-ackwards: The development of
hierarchical factor structures from the top down. *Journal of Research in
Personality*, *40*(4), 347–358. https://doi.org/10.1016/j.jrp.2006.01.001

Waller, N. G. (2007). A general method for computing hierarchical component
structures by *Bass-Ackward* factor analysis. *Journal of Research in
Personality*, *41*(4), 745–752. https://doi.org/10.1016/j.jrp.2006.08.005

Horn, J. L. (1965). A rationale and test for the number of factors in factor
analysis. *Psychometrika*, *30*(2), 179–185.

Velicer, W. F. (1976). Determining the number of components from the matrix of
partial correlations. *Psychometrika*, *41*(3), 321–327.
