---
title: "Introduction to Bass-Ackwards Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Bass-Ackwards Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

## The problem: which factor solution is right?

Anyone who has applied factor analysis faces the same dilemma: how many factors
should you extract? Parallel analysis might say 5; the scree plot might suggest
3; a reviewer might insist on 1. Each solution seems to describe the data
differently, and it is tempting to treat them as competing answers to the same
question.

Goldberg (2006) reframed this problem: the different solutions are not competing
— they are complementary. A 1-factor solution captures the broadest shared
variance in the data, the dimension along which all variables correlate. A
5-factor solution captures narrower, more specific dimensions. These two levels
of resolution describe the same underlying structure from different vantage
points, just as a satellite image and a street map describe the same city.

The **bass-ackwards method** makes this relationship explicit. It fits factor
models at every level from 1 up to k, and then computes the correlations between
factor *scores* across adjacent levels. Those correlations show you:

- **Which narrow factor inherits from which broad factor** — the lineage of
  each dimension as you move from coarse to fine resolution.
- **Where a factor splits** — a single broad factor that correlates strongly
  with two narrower factors is fragmenting into sub-dimensions.
- **Where a factor is stable** — a factor that correlates ~1.0 with its
  counterpart at the adjacent level is essentially the same construct appearing
  at two resolutions.

The result is a **hierarchy map** — a visual and numerical account of how
the factor structure of your data builds from broad to narrow.

> *Note: This is a descriptive, data-driven hierarchy, not a confirmatory
> hierarchical model like Schmid–Leiman or higher-order SEM. The between-level
> correlations are score correlations (or their algebraic equivalents), not
> model parameters.*

## Data

We use `bfi25`, the built-in 25-item Big Five example dataset.
The items measure five personality traits — Agreeableness (A1–A5),
Conscientiousness (C1–C5), Extraversion (E1–E5), Neuroticism (N1–N5), and
Openness (O1–O5) — each on a 6-point Likert scale. See `?bfi25` for full
provenance; the data derive from the SAPA project using public-domain IPIP items.

```{r data}
library(ackwards)

bfi <- na.omit(bfi25)
dim(bfi)
```

We drop cases with any missing item. In a real analysis you might use pairwise
deletion or multiple imputation; `na.omit()` is sufficient for this illustration.

## Step 1: How many factors? `suggest_k()`

Before fitting the hierarchy, get a sense of the plausible range of k.
`suggest_k()` runs five complementary selection criteria and reports a consensus
range:

```{r suggest_k}
sk <- suggest_k(bfi, seed = 42)
print(sk)
```

```{r suggest_k_plot, fig.width=6, fig.height=7}
autoplot(sk)
```

`suggest_k()` does not return a single "correct" k — it reports what several
criteria each recommend, so you can read a plausible *range* rather than a point
estimate. That range informs `k_max` in `ackwards()`, which is itself an *upper
bound* on the hierarchy depth, not a claim about the true number of factors:
setting `k_max` one or two levels above the consensus to watch factors fragment
is intentional and informative. For a full explanation of each criterion, its
bias direction, and how to match it to your engine, see
`vignette("ackwards-suggest-k")`.

## Step 2: Fit the hierarchy `ackwards()`

Now fit the bass-ackwards hierarchy. The most important arguments are:

| Argument | What it controls | Default |
|----------|-----------------|---------|
| `k_max` | Maximum depth of the hierarchy | *(required)* |
| `engine` | Extraction engine: `"pca"`, `"efa"`, or `"esem"` | `"pca"` |
| `cor` | Correlation type: `"pearson"`, `"spearman"`, `"polychoric"` | `"pearson"` |

Because BFI items are ordinal, we use `cor = "polychoric"`. This computes
polychoric correlations between items before factor extraction, giving a more
accurate representation of the latent structure.

```{r fit}
x <- ackwards(bfi, k_max = 5, cor = "polychoric")
```

No warning this time — specifying `cor = "polychoric"` tells `ackwards()` that
you have already thought about the measurement scale.

> *Why varimax? Within each level, `ackwards()` rotates the factors orthogonally
> using varimax, and this is not merely a cosmetic default. Orthogonality is what
> gives the between-level score correlations a closed form: an orthonormal
> rotation satisfies T′ = T⁻¹, which is exactly what makes the `W′RW` edge algebra
> exact rather than an approximation. Varimax is the orthogonal rotation Goldberg
> (2006) used, and it is the same rotation as the "CF-VARIMAX" reported by
> Mplus-based papers such as Kim & Eaton (2015) — `CF(κ = 1/p)` is varimax — so
> the two labels are not competing choices. Oblique rotations are deliberately not
> offered: correlated factors would confound the cross-level signal the method
> exists to measure.*

## Step 3: Summarize the result

### High-level summary

`print()` gives a quick overview:

```{r print}
print(x)
```

The "Levels" section confirms that all five models converged, and reports the
cumulative variance explained at each level. Notice that the jump from k = 1 to
k = 2 is large (`r sprintf("%.1f%%", 100 * x$levels[["1"]]$variance[["cumulative"]])`
→ `r sprintf("%.1f%%", 100 * x$levels[["2"]]$variance[["cumulative"]])`), while
later jumps are smaller — characteristic of data with a strong general factor
and several specific dimensions.

The "Edges" section reports how many of the 40 possible between-level
connections exceed the display threshold of |r| ≥ 0.3. The 40 comes from
summing across all adjacent pairs: 1×2 + 2×3 + 3×4 + 4×5 = 2 + 6 + 12 + 20 =
40.

`summary()` gives a more detailed view — per-factor variance and fit indices at
each level, plus a lineage list showing which factors at each level descend from
which parents:

```{r summary}
summary(x)
```

`glance()` returns the same top-level information as a one-row data frame, convenient for
comparisons across models:

```{r glance}
glance(x)
```

## Step 4: Visualize the hierarchy `autoplot()`

The hierarchy diagram is the centerpiece of the method. Each row represents one
level (k = 1 at the top, k = 5 at the bottom). Arrows connect each factor to its
**primary parent** — the factor at the level above with which it has the
strongest correlation (|r|). Arrow thickness is proportional to |r|, and color
shows direction (blue = positive, red = negative by default); both aesthetics
come with a legend. For a left-to-right layout (handy for wide slides), pass
`direction = "horizontal"`.

```{r autoplot, fig.width = 8, fig.height = 5, warning = FALSE}
autoplot(x)
```

Reading this diagram from broad (top) to narrow (bottom) tells the story of the
Big Five:

- **k = 1**: One broad factor. Given the polychoric basis, this captures shared
  variance across all 25 items.
- **k = 2**: The broad factor splits into two: Neuroticism separates from a
  mixed cluster of Extraversion, Agreeableness, Conscientiousness, and
  Openness items (Extraversion items load most strongly within that cluster).
- **k = 3**: The non-Neuroticism cluster splits again, separating
  Extraversion/Agreeableness from Conscientiousness/Openness; Neuroticism
  stays intact and sheds its residual cross-loadings.
- **k = 4**: Conscientiousness and Openness differentiate into their own
  factors.
- **k = 5**: Extraversion and Agreeableness finally differentiate, completing
  the canonical Big Five.

The factors are labeled `m{k}f{j}` (level k, factor j). These stable IDs are
used throughout the object, so `m5f1` at k = 5 refers to the same factor in
the loadings, the edge table, the factor scores, and the diagram.

### Adjusting the diagram

`autoplot()` accepts arguments to control edge thresholds, colours, line
styles, arrowheads, node labels, and more. See
`vignette("ackwards-visualization")` for a guided tour of all options with
rendered examples, or `?autoplot.ackwards` for the full argument list.

## Step 5: Dig into the factors `tidy()`

`tidy()` extracts any component of the result as a tidy data frame. The `what`
argument controls what is returned.

### Reading each factor — `top_items()`

To understand what each factor represents, `top_items()` lists the salient
items (those with `|loading| >= cut`) for every factor, grouped by level.
This is more readable than a full item-by-factor matrix, especially for deep
hierarchies.

```{r top-items}
top_items(x, level = 5, cut = 0.5)
```

Reading a hierarchy of factors and giving them names is its own topic — including
the sign convention, naming across levels, and applying labels to the diagram.
See `vignette("ackwards-interpret")` for the full workflow.

### Factor loadings (tidy)

For programmatic access, `tidy(what = "loadings")` returns the full loading
matrix in long format — one row per item × factor × level:

```{r tidy-loadings}
loadings_df <- tidy(x, what = "loadings")
head(loadings_df)
```

The `se`, `ci_lower`, and `ci_upper` columns are `NA` here. PCA (and EFA) return
point loadings with no standard errors, so there is nothing to report. These
columns are populated only by `engine = "esem"`, which fits each level in lavaan
and can attach model-based standard errors and confidence intervals to the
loadings; see `vignette("ackwards-engines")`.

### Between-level edges

Each edge is the between-level factor-score correlation computed via Waller's
(2007) closed-form W′RW algebra — an exact result that requires no score
materialization, just the weight matrices and the input correlation matrix.

```{r tidy-edges}
# Each factor's primary-parent edge, strongest first
tidy(x, what = "edges", primary_only = TRUE, sort = "strength")
```

`primary_only = TRUE` keeps just the strongest-connecting edge for each factor
— its primary parent — and `sort = "strength"` orders them by `|r|`. The r
values close to 1.0 indicate factors that are nearly identical across adjacent
levels — a sign of a stable, replicable dimension. Smaller values indicate where
the structure is reorganizing.

### Variance explained

```{r tidy-variance}
tidy(x, what = "variance")
```

Each row is one factor at one level. `proportion` is the fraction (0-1) of total
item variance explained by that factor; `cumulative` accumulates within a
level. Multiply by 100 for a percentage.

## Step 6: Score observations `augment()`

Factor scores place every observation on each factor at every level. They are
useful for regression, clustering, or any downstream analysis where you want a
continuous summary of a latent dimension.

`augment(x, data = bfi)` computes scores on the fly from the stored weight
matrices. By default it appends them to your data frame; pass `append = FALSE` to
get just the score columns, named `.m{k}f{j}`. Row order is preserved either way,
so the scores line up with the input rows.

```{r augment}
scores <- augment(x, data = bfi, append = FALSE)
dim(scores) # 15 score columns (1+2+3+4+5)
names(scores)
```

Scores are standardized so that the **model-implied** variance is 1 — meaning
the scaling comes from the polychoric correlation matrix, not from the raw data.
In practice the empirical standard deviations will be close to but not exactly 1
when a non-Pearson basis is used; see `?augment.ackwards` for details. For pure
PCA on Pearson correlations the model-implied and empirical variances agree
exactly.

### Using scores for downstream analysis

Because `augment()` returns a plain data frame, the scores slot directly into
any standard R workflow. As a quick illustration, the k = 5 factor scores should
be nearly uncorrelated with each other — a consequence of orthogonal rotation —
while scores across levels should be highly correlated along the primary-parent
lineage:

```{r downstream}
# Within-level correlations at k = 5: should be near zero (orthogonal rotation)
k5 <- scores[, c(".m5f1", ".m5f2", ".m5f3", ".m5f4", ".m5f5")]
round(cor(k5), 2)
```

```{r downstream2}
# Cross-level: m4f1 is the primary parent of m5f1 and m5f4 (from the edge table)
lineage <- scores[, c(".m4f1", ".m5f1", ".m5f2", ".m5f4")]
round(cor(lineage), 2)
```

The cross-level block confirms lineage: `.m4f1` correlates strongly with
`.m5f1` and `.m5f4` (the k = 5 factors it spawned) and near-zero with `.m5f2`
(which descends from a different k = 4 factor). This is a sanity check you can
run on any result — strong parent–child correlations should appear exactly where
the `tidy(what = "edges")` table says they should.

## Summary

The bass-ackwards workflow in **ackwards** has six steps:

1. **`suggest_k(data)`** — identify a plausible range for the hierarchy depth.
2. **`ackwards(data, k_max, cor = ...)`** — fit the full hierarchy of factor models.
3. **`print(x)` / `summary(x)` / `glance(x)`** — check convergence, read the
   per-level variance and fit indices, and inspect the lineage list.
4. **`autoplot(x)`** — visualize the hierarchy as a lineage diagram.
5. **`tidy(x, what = ...)`** — extract loadings, edges, or variance in a form
   ready for tables or further analysis.
6. **`augment(x, data = ...)`** — generate factor scores for downstream use.

## Next steps

| Topic | Vignette |
|-------|---------|
| Choosing k: the five criteria in depth, pros/cons, and best practices | `vignette("ackwards-suggest-k")` |
| When PCA is not enough: comparing EFA and ESEM engines | `vignette("ackwards-engines")` |
| Ordinal data: polychoric correlations and WLSMV estimation | `vignette("ackwards-ordinal")` |
| Skip-level connections and pruning with the Forbes extension | `vignette("ackwards-forbes")` |
| Customizing the hierarchy diagram | `vignette("ackwards-visualization")` |

## References

Goldberg, L. R. (2006). Doing it all bass-ackwards: The development of
hierarchical factor structures from the top down. *Journal of Research in
Personality*, *40*(4), 347–358. https://doi.org/10.1016/j.jrp.2006.01.001

Horn, J. L. (1965). A rationale and test for the number of factors in factor
analysis. *Psychometrika*, *30*(2), 179–185.

Velicer, W. F. (1976). Determining the number of components from the matrix of
partial correlations. *Psychometrika*, *41*(3), 321–327.

Waller, N. G. (2007). A general method for computing hierarchical component
structures by *Bass-Ackward* factor analysis. *Journal of Research in
Personality*, *41*(4), 745–752. https://doi.org/10.1016/j.jrp.2006.08.005
