---
title: "The Forbes Extension: Skip-Level Connections and Pruning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Forbes Extension: Skip-Level Connections and Pruning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

The classic Goldberg (2006) method computes between-level factor-score
correlations only for *adjacent* levels: 1↔2, 2↔3, 3↔4, and so on. Forbes
(2023) extended this in two ways: computing correlations across *all* level
pairs (not just adjacent), and using those extra connections to identify and
flag **redundant** or **artefactual** factors in the hierarchy.

This vignette covers both extensions: `pairs = "all"` and `prune`.

## The limitation of adjacent-only edges

An adjacent-only hierarchy shows you the immediate parent–child relationships.
What it cannot tell you is whether a factor at level k is essentially the same
construct as a factor several levels up — a sign that the intermediate levels
are adding noise rather than resolution.

Consider a factor that appears at k = 2, k = 3, and k = 4 and correlates
> 0.97 with its counterpart at every adjacent level. The adjacent-only diagram
shows three consecutive arrows, each nearly perfect. But nothing in the diagram
directly flags that the k = 3 factor is redundant: you could skip straight from
k = 2 to k = 4 without losing any information.

**Skip-level correlations** make this visible by computing the correlation
between every pair of levels, not just neighbors.

## Setup

```{r setup-data}
library(ackwards)
bfi <- na.omit(psych::bfi[, 1:25])
```

## `pairs = "all"`: computing every between-level correlation

Adding `pairs = "all"` extends the edge table from adjacent pairs only to every
combination of levels.

```{r adj-vs-all}
# Classic adjacent-only
x_adj <- ackwards(bfi, k = 5, cor = "polychoric")

# All pairs
x_all <- ackwards(bfi, k = 5, cor = "polychoric", pairs = "all")

# How many edges?
nrow(tidy(x_adj, what = "edges")) # adjacent only
nrow(tidy(x_all, what = "edges")) # all pairs
```

With k = 5, the adjacent-only model has 40 edges (1×2 + 2×3 + 3×4 + 4×5).
The all-pairs model adds every non-adjacent pair — 1↔3, 1↔4, 1↔5, 2↔4, 2↔5,
3↔5 — for 85 edges total.

### Reading the skip-level edge table

```{r skip-edges}
edges <- tidy(x_all, what = "edges")

# Non-adjacent edges with |r| >= 0.5, sorted by strength
skip <- edges[abs(edges$level_to - edges$level_from) > 1 & abs(edges$r) >= 0.5, ]
skip <- skip[order(-abs(skip$r)), c("from", "to", "level_from", "level_to", "r")]
head(skip, 12)
```

Several factors connect across two or more levels with correlations above 0.90.
`m3f2` (level 3, factor 2) correlates 0.98 with `m5f1` (level 5, factor 1),
jumping *two* levels. This tells you that m3f2 and m5f1 are essentially the
same construct — the intermediate levels are just refinements within a stable
dimension.

## Pruning: identifying redundant factors

The `prune` argument uses the skip-level correlations to automatically flag
factors that may not be adding genuine information to the hierarchy.

### `prune = "redundant"`: chains of near-identical factors

A **redundant chain** is a sequence of factors connected by near-perfect
correlations (|r| ≥ 0.9 by default) across levels. If m2f2 → m3f2 → m4f2 →
m5f1 all share r > 0.97, the intermediate nodes m3f2 and m4f2 are flagged as
redundant: they repeat rather than refine the same dimension.

```{r prune-redundant}
x_prune <- ackwards(bfi,
  k = 5, cor = "polychoric",
  pairs = "all", prune = "redundant"
)
```

```{r prune-nodes}
tidy(x_prune, what = "nodes")
```

Six factors are flagged as redundant. The entire k = 4 level (m4f1 through
m4f4) is flagged, as are m2f2 and m3f2. This is a striking finding: for this
dataset and this k, the four-factor level adds little beyond what you already
know from k = 3 and k = 5.

The flagged factors are not removed from the object — `prune` is purely a
diagnostic annotation, not a deletion. You can still inspect their loadings, use
their scores, and include them in the diagram. Pruning flags guide interpretation;
they do not alter the model.

### The pruned-factor diagram

For presentations and publications it is cleaner to omit the flagged factors
entirely and draw direct connections from each retained factor to its single
strongest kept ancestor — even when that ancestor is several levels away. This
is the Forbes (2023) pruned-factor diagram, activated by `drop_pruned = TRUE`.

Correlation labels appear by default in this view (`show_r` auto-defaults to
`TRUE`) because they are the primary signal distinguishing a close connection
from a long gap-spanning one. The arguments below reproduce the style of the
figures in Forbes (2023): black lines, uniform width, plain line ends, and no
legend.

```{r drop-pruned, warning = FALSE}
autoplot(x_prune,
  drop_pruned = TRUE,
  color_pos = "black", color_neg = "black",
  edge_linewidth = 0.6, show_arrows = FALSE, legend = FALSE
)
```

Level 4 is entirely pruned, leaving a visible gap in the y-axis. The gap is
intentional: it shows *which* level was removed. Spanning arrows bridge
directly from level 3 factors to level 5 factors (and from level 2 to level 5
where intermediate levels were flagged). Solid lines indicate strong connections
(|r| ≥ 0.5 by default); dashed lines indicate weaker ones — the same
`cut_strong` threshold used throughout.

To close the gaps and compact the layout while retaining the original level
numbers on the axis:

```{r drop-pruned-compressed, warning = FALSE}
autoplot(x_prune,
  drop_pruned = TRUE, compress_levels = TRUE,
  color_pos = "black", color_neg = "black",
  edge_linewidth = 0.6, show_arrows = FALSE, legend = FALSE
)
```

The level labels still read "1 factor", "2 factors", "3 factors", "5 factors"
so readers know which levels were retained; the uniform vertical spacing makes
the diagram easier to read in constrained page layouts.

For further cosmetic customization — colours, node labels, arrowheads, and
more — see `vignette("ackwards-visualization")`.

### `prune = "artefact"`: factors defined by structural similarity

An **artefact** factor is one whose loading pattern is more similar to a factor
at a *non-adjacent* level than to its own-level neighbors. Similarity is
measured by Tucker's congruence coefficient (φ):

$$\phi(F_a, F_b) = \frac{\sum_i \lambda_{ia}\lambda_{ib}}
{\sqrt{\sum_i \lambda_{ia}^2 \cdot \sum_i \lambda_{ib}^2}}$$

φ ranges from −1 to +1, with values > 0.95 indicating near-identical loading
patterns regardless of sign. A factor at k = 3 that has φ > 0.95 with a factor
at k = 1 is suspected to be an artefact of the rotation rather than a genuine
new dimension.

```{r prune-artefact}
x_art <- ackwards(bfi,
  k = 5, cor = "polychoric",
  pairs = "all", prune = "artefact"
)
tidy(x_art, what = "nodes")
```

For the BFI, the artefact criterion flags nothing — no factor at any level has a
loading pattern more similar to a factor from a non-adjacent level than to its
own-level solution. This is a good result for a well-validated instrument: it
suggests the rotation is producing genuinely distinct factors at each level,
not recycling old ones.

On data with weaker or noisier factor structure, or with rotations that struggle
to separate highly correlated factors, the artefact criterion will often flag
some nodes. A node can be flagged by one criterion but not the other; the two
criteria capture different flavors of redundancy:

- `"redundant"`: this factor appears at multiple levels with near-identical
  *score correlations* — it persists unchanged as k increases.
- `"artefact"`: this factor's *loading pattern* closely resembles a factor from
  a different, non-adjacent level — it looks like a copy rather than a
  refinement.

## Tuning the thresholds

The redundancy criterion has an adjustable `redundancy_r` threshold (default
`0.90`) matching Forbes (2023). The artefact criterion has no auto-flag
threshold — `prune = "artefact"` computes Tucker's φ for researcher inspection;
no factors are auto-flagged.

For the BFI, the result is the same across a wide range of thresholds because
the redundant chains all have correlations > 0.97 — the flagging is unambiguous.
With your own data you may find borderline cases where the threshold matters:

```{r thresholds}
# Check how many factors are flagged at different redundancy_r thresholds
thresholds <- c(0.80, 0.85, 0.90, 0.95)
counts <- sapply(thresholds, function(thr) {
  x <- suppressWarnings(
    ackwards(bfi,
      k = 5, cor = "polychoric",
      pairs = "all", prune = "redundant",
      redundancy_r = thr
    )
  )
  sum(tidy(x, what = "nodes")$pruned)
})
data.frame(redundancy_r = thresholds, n_flagged = counts)
```

For the BFI all thresholds agree: the flagged factors are robustly redundant, not
borderline cases. In noisier datasets or smaller samples you will typically see
the count increase as you lower the threshold.

## Practical interpretation

The Forbes extension does not change the core bass-ackwards analysis. It
enriches it with two questions:

1. **Do any factors persist unchanged across multiple levels?** (`pairs = "all"`)
   Skip-level correlations near 1.0 indicate stable dimensions that survive
   changes in k — exactly the kind of robust construct you want to report.

2. **Are there levels where the factor structure is just reorganizing rather
   than genuinely differentiating?** (`prune = "redundant"`)
   Flagged levels can often be removed from the k range without losing
   interpretive content.

A common workflow: fit with `pairs = "all"` first to examine the full picture,
then apply `prune = "redundant"` to identify which levels add the most new
information, and use that to guide your focus in reporting.

## References

Goldberg, L. R. (2006). Doing it all bass-ackwards. *Journal of Research in
Personality*, *40*(4), 347–358.

Forbes, M. K. (2023). Improving hierarchical models of individual differences:
An extension of Goldberg's bass-ackward method. *Psychological Methods*.
https://doi.org/10.1037/met0000593

Tucker, L. R. (1951). *A method for synthesis of factor analysis studies*
(Personnel Research Section Report No. 984). Department of the Army.
