---
title: "Customizing the Hierarchy Diagram"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customizing the Hierarchy Diagram}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE
)
```

`autoplot.ackwards()` exposes a large number of arguments for controlling the
appearance of the hierarchy diagram. This vignette is a visual reference: each
section demonstrates one group of arguments with rendered figures so you can
see the effect before writing any code.

All options shown here are **presentation-only** — they do not change which
factors were extracted or how the between-level correlations were computed.
Options that change *which nodes appear* (`drop_pruned`, `compress_levels`) are
specific to the Forbes pruning extension and are covered in
`vignette("ackwards-forbes")`.

## Setup

```{r setup-data}
library(ackwards)

bfi <- na.omit(psych::bfi[, 1:25])
x <- ackwards(bfi, k = 5, cor = "polychoric")
```

The default diagram for reference:

```{r base}
autoplot(x)
```

Factors are labeled `m{k}f{j}` (level k, factor j). Arrow **thickness** encodes
|r|; **colour** encodes direction (blue = positive, red–orange = negative);
**linetype** encodes strength (solid: |r| ≥ `cut_strong`; dashed: `cut_show` ≤
|r| < `cut_strong`). Level labels on the left count factors per level.

## Filtering edges

### `cut_show` — minimum |r| to display

Edges below `cut_show` are hidden entirely. Raising it produces a sparser
diagram that emphasises only the strongest connections.

```{r cut-show}
# Default cut_show = 0.3 (already shown above)
autoplot(x, cut_show = 0.5)
```

### `cut_strong` — solid vs dashed threshold

Edges at or above `cut_strong` are drawn solid; edges below it (but above
`cut_show`) are dashed. Raising `cut_strong` makes only the very strongest
connections solid.

```{r cut-strong}
autoplot(x, cut_show = 0.3, cut_strong = 0.8)
```

## Edge colours

### `color_pos` / `color_neg` — custom direction colours

The default blue/red palette can be replaced with any colours recognised by R.

```{r colours}
autoplot(x, color_pos = "darkorchid", color_neg = "darkorange")
```

Setting both to the same colour produces uniformly coloured edges while
retaining the linetype-based strength distinction from `cut_strong`. This is
the basis for the Forbes (2023) publication style (see the worked example at the
end of this vignette).

## Monochrome mode

### `mono = TRUE` — encode direction on linetype

`mono = TRUE` switches the direction encoding from colour to linetype: solid
lines are positive correlations and dashed lines are negative correlations.
`linewidth` still encodes |r|. The `cut_strong` strength distinction is dropped
in this mode because linewidth already conveys magnitude.

```{r mono}
autoplot(x, mono = TRUE)
```

`mono` is suited for black-and-white figures where the reader must distinguish
positive from negative edges. Note that the linetype semantics differ from the
colour-mode dashing: here dashed = *negative sign*, not *weak connection*. For
the Forbes-style uniform-line look (dashed = weak, regardless of sign), use
`color_pos = color_neg = "black"` in colour mode instead — see the worked
example below.

Combining `mono` with `show_r` labels the edges with their exact values,
removing any ambiguity about magnitude:

```{r mono-show-r}
autoplot(x, mono = TRUE, show_r = TRUE)
```

## Correlation labels

### `show_r` / `r_digits` — annotate edges with |r| values

`show_r = TRUE` draws the rounded correlation at each edge midpoint.
`r_digits` controls the number of decimal places (default 2).

```{r show-r}
autoplot(x, show_r = TRUE)
```

```{r show-r-digits}
autoplot(x, show_r = TRUE, r_digits = 1L)
```

## Node labels

### `node_labels` — rename individual factors

`node_labels` is a named character vector mapping factor IDs to display strings.
Unspecified factors keep their `m{k}f{j}` labels.

```{r node-labels}
autoplot(x, node_labels = c(
  m5f1 = "Neuro.",
  m5f2 = "Extra.",
  m5f3 = "Open.",
  m5f4 = "Agree.",
  m5f5 = "Consc."
))
```

Multi-line labels are supported via `\n`:

```{r node-labels-multiline}
autoplot(x, node_labels = c(
  m5f1 = "Neuro-\nticism",
  m5f2 = "Extra-\nversion"
))
```

## Structural simplifications

### `primary_only = TRUE` — show only primary-parent edges

Setting `primary_only = TRUE` keeps only the single strongest edge per factor
(its primary parent), producing a clean tree. Because skip-level edges are
never primary, this also suppresses curved arcs when `pairs = "all"` was used.

```{r primary-only}
autoplot(x, primary_only = TRUE)
```

## Level labels

### `show_level_labels` / `level_label_size`

Level labels ("1 factor", "2 factors", …) are shown by default on the left
margin. They can be hidden or resized.

```{r level-labels-off}
autoplot(x, show_level_labels = FALSE)
```

```{r level-labels-size}
autoplot(x, show_level_labels = TRUE, level_label_size = 4)
```

## Arrowheads

### `show_arrows = FALSE` — plain line ends

By default, edges end with closed arrowheads. Setting `show_arrows = FALSE`
draws plain line ends. This applies to both straight edges and curved skip-level
arcs.

```{r no-arrows}
autoplot(x, show_arrows = FALSE)
```

## Edge width

### `edge_linewidth` — uniform vs. |r|-scaled width

By default, edge width is proportional to |r|. A numeric `edge_linewidth` draws
every edge at that constant width and removes the linewidth legend.

```{r edge-linewidth}
autoplot(x, edge_linewidth = 0.7)
```

## Legend

### `legend = FALSE` — suppress all guides

`legend = FALSE` removes all legends from the plot. Most useful when the diagram
is self-explanatory or the legend duplicates information conveyed by labels.

```{r no-legend}
autoplot(x, legend = FALSE)
```

## Worked example: publication-ready figure

The following call reproduces the visual style of Forbes (2023): black lines of
uniform weight, plain line ends, correlation labels, and no legend. It uses
colour mode (not `mono`) so that the dashing retains its `cut_strong`
weak/secondary-connection semantics.

```{r pub-figure}
autoplot(x,
  color_pos      = "black",
  color_neg      = "black",
  edge_linewidth = 0.6,
  show_arrows    = FALSE,
  show_r         = TRUE,
  legend         = FALSE
)
```

Combining with `node_labels` names the factors for the final figure:

```{r pub-figure-labeled}
autoplot(x,
  color_pos = "black",
  color_neg = "black",
  edge_linewidth = 0.6,
  show_arrows = FALSE,
  show_r = TRUE,
  legend = FALSE,
  node_labels = c(
    m5f1 = "Neuro.",
    m5f2 = "Extra.",
    m5f3 = "Open.",
    m5f4 = "Agree.",
    m5f5 = "Consc."
  )
)
```

For the pruned-factor variant of this figure (nodes omitted, spanning arrows)
see `vignette("ackwards-forbes")`.
