---
title: "Ordinal Data: Polychoric Correlations and WLSMV"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ordinal Data: Polychoric Correlations and WLSMV}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
has_lavaan <- requireNamespace("lavaan", quietly = TRUE)
```

Most psychological scales use ordinal response formats — Likert-type items with
4, 5, 6, or 7 response options. Treating these as continuous and computing
Pearson correlations between them is ubiquitous in practice, but it introduces
systematic bias that can distort both loadings and the between-level edges that
bass-ackwards analysis depends on. This vignette explains the problem and how
`ackwards` addresses it.

## Why Pearson underestimates for ordinal items

When a continuous latent trait is sliced into a small number of ordered
categories, the observed correlation between two items is always lower than the
correlation between their underlying latent variables. The more coarse the
rating scale, the larger the attenuation. For a 5-point scale, Pearson
correlations can be 10–20% lower than the true latent correlations; for a
2-point (binary) scale, the underestimate can exceed 30%.

This matters for bass-ackwards analysis because:

1. **Loadings are attenuated.** Factors appear weaker than they are.
2. **Between-level edges are attenuated.** The hierarchy looks flatter —
   factors that should be nearly perfectly correlated across levels may appear
   only moderately correlated.
3. **The number of factors is underestimated.** Attenuation compresses the
   eigenvalue spectrum, so selection criteria (parallel analysis, MAP) may
   suggest fewer factors than truly exist.

## The polychoric solution

A **polychoric correlation** between two ordinal items estimates the Pearson
correlation that *would* exist between their underlying continuous latent
variables, if those variables had been measured without discretization. It does
this by fitting a bivariate normal model with thresholds at each category
boundary.

Setting `cor = "polychoric"` in `ackwards()` replaces step 1 of the pipeline —
computing the item correlation matrix `R` — with a polychoric matrix. All
downstream computation (factor extraction, rotation, tenBerge scoring,
between-level algebra) then operates on that polychoric `R`.

## Automatic detection

`ackwards()` checks for ordinal-looking data using a simple heuristic: if any
column is integer-valued with ≤ 7 distinct levels, the data is flagged and a
warning is issued when `cor = "pearson"` (the default).

```{r data}
library(ackwards)
bfi <- na.omit(psych::bfi[, 1:25])
```

```{r detect}
# The warning fires with the default Pearson basis
x_pearson <- ackwards(bfi, k_max = 5)
```

The warning is ackwards reminding you to make an active choice. Specifying
`cor = "polychoric"` acknowledges the data type and suppresses it:

```{r polychoric-silent}
x_poly <- ackwards(bfi, k_max = 5, cor = "polychoric")
```

## Seeing the difference

### Correlation matrices

The most direct way to see the impact of the basis is to compare the
correlation matrices directly.

```{r r-compare}
# Pearson R (stored in x$r)
round(x_pearson$r[1:5, 1:5], 2)
```

```{r r-compare2}
# Polychoric R
round(x_poly$r[1:5, 1:5], 2)
```

The polychoric correlations are consistently higher — sometimes substantially
so. The `N1`–`N2` pair, for instance, goes from roughly 0.59 (Pearson) to 0.73
(polychoric): a difference large enough to shift loadings and change
eigenvalues.

### Loadings

```{r load-compare}
l_pear <- tidy(x_pearson, what = "loadings")
l_poly <- tidy(x_poly, what = "loadings")
l_pear$basis <- "pearson"
l_poly$basis <- "polychoric"

# k = 5 loadings for Neuroticism items
both <- rbind(l_pear, l_poly)
both5_N <- both[both$level == 5 & substr(both$item, 1, 1) == "N", ]
both5_N[
  order(both5_N$factor, both5_N$item),
  c("basis", "factor", "item", "loading")
]
```

Polychoric loadings for the Neuroticism items are noticeably higher — the latent
structure is sharper when the attenuating effect of coarse categories is
removed.

### Between-level edges

```{r edge-compare}
e_pear <- tidy(x_pearson, what = "edges")
e_poly <- tidy(x_poly, what = "edges")

primary_pear <- e_pear[e_pear$is_primary, c("from", "to", "r")]
primary_poly <- e_poly[e_poly$is_primary, c("from", "to", "r")]

primary_pear$basis <- "pearson"
primary_poly$basis <- "polychoric"

both_e <- rbind(primary_pear, primary_poly)
both_e[order(both_e$to, both_e$basis), ]
```

The edges are broadly similar in sign and rank order — the hierarchy is the
same — but polychoric edges are stronger. Factors that represent genuinely stable
dimensions show near-perfect correlations with their counterparts at adjacent
levels when the true latent structure is properly recovered.

## WLSMV for ESEM with ordinal items

When `engine = "esem"` and `cor = "polychoric"` are combined, `ackwards()`
automatically switches the lavaan estimator to **WLSMV** (diagonally weighted
least squares, mean- and variance-adjusted). This is the standard estimator for
structural equation models with categorical or ordinal indicators, and is the
same method used by Mplus by default.

```{r esem-check, echo = FALSE, results = "asis"}
if (!has_lavaan) {
  cat(
    "> **Note:** The ESEM examples below require the **lavaan** package,",
    "which is not currently installed. Install it with",
    "`install.packages(\"lavaan\")` to run these chunks.\n"
  )
}
```

```{r esem-wlsmv, eval = has_lavaan}
x_esem <- ackwards(bfi, k_max = 3, engine = "esem", cor = "polychoric")
x_esem
```

The WLSMV fit indices (CFI, RMSEA, SRMR) are now computed on a model that
respects the ordinal measurement level. They will typically look better than
Pearson-based fit indices on the same data, reflecting both the better model
specification and the different reference distribution WLSMV uses.

```{r esem-fit, eval = has_lavaan}
tidy(x_esem, what = "fit")
```

## Practical guidance: when to use polychoric

| Scale characteristics | Recommendation |
|-----------------------|---------------|
| Continuous or many categories (> 7) | `cor = "pearson"` (default) |
| 5–7 ordered categories (typical Likert) | `cor = "polychoric"` — recommended |
| 3–4 categories | `cor = "polychoric"` — strongly recommended |
| Binary (2 categories) | `cor = "polychoric"` — essential; consider tetrachoric specifically |

Polychoric estimation requires fitting a bivariate normal model for every item
pair. With p items that is p(p−1)/2 pairs — about 300 for a 25-item scale.
Computation is usually fast (seconds), but scales as O(p²) and can be slow for
very large item pools.

> **A note on score computation.** When `cor = "polychoric"`, the weight
> matrices stored in the object are derived from the polychoric `R`. Materializing
> factor scores via `augment()` applies those weights to Pearson-standardized raw
> data (`scale(data)`). The resulting scores are calibrated to have
> *model-implied* unit variance, not empirically unit variance. For downstream
> analysis this distinction rarely matters — the scores are still well-scaled —
> but it is why `augment()` issues a warning when `cor != "pearson"`. The
> between-level edges from `tidy(what = "edges")` are always exact because they
> come from the algebra, not materialized scores.

## References

Pearson, K. (1900). Mathematical contributions to the theory of evolution. VII.
On the correlation of characters not quantitatively measurable. *Philosophical
Transactions of the Royal Society of London, Series A*, *195*, 1–47.

Olsson, U. (1979). Maximum likelihood estimation of the polychoric correlation
coefficient. *Psychometrika*, *44*(4), 443–460.

Muthén, B. O. (1984). A general structural equation model with dichotomous,
ordered categorical, and continuous latent variable indicators.
*Psychometrika*, *49*(1), 115–132.

Flora, D. B., & Curran, P. J. (2004). An empirical evaluation of alternative
methods of estimation for confirmatory factor analysis with ordinal data.
*Psychological Methods*, *9*(4), 466–491.
