---
title: "Interpreting and Labeling Factors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpreting and Labeling Factors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 7
)
```

```{r packages, message = FALSE, warning = FALSE}
library(ackwards)
bfi <- na.omit(psych::bfi[, 1:25])
x <- ackwards(bfi, k_max = 5)
```

Fitting an `ackwards` model gives you a hierarchy of factors with stable but
opaque IDs — `m1f1`, `m2f1`, `m2f2`, and so on. Turning those IDs into something
you can reason about is *interpretive* work: you read each factor's loadings,
decide what construct it represents, and give it a name. This is the part of the
analysis that requires judgment, and it is harder in a bass-ackwards hierarchy
than in a single flat factor solution, because the same construct can appear,
split, and merge across levels.

This article covers the full workflow:

1. **Read** each factor with `top_items()`.
2. **Understand** the sign convention so you don't misread a factor.
3. **Name** factors in a way that respects the hierarchy.
4. **Apply** your names to the diagram with `label_template()` and
   `autoplot(node_labels = ...)`.

## Reading a factor with `top_items()`

A factor is defined by the items that load strongly on it. `top_items()` lists,
for each factor, the items whose absolute loading meets a threshold, sorted from
strongest to weakest. This is far easier to read than a full item-by-factor
matrix, especially at deeper levels.

```{r top-items-5}
top_items(x, level = 5)
```

The default `cut = 0.3` shows every item with `|loading| >= 0.3`. Raise it to
isolate the items that most define each factor, or lower it to surface weaker
cross-loadings.

```{r top-items-cut}
top_items(x, level = 5, cut = 0.45)
```

When a factor has many salient items, `n` caps the list at the strongest few so
you can see the core of each factor at a glance:

```{r top-items-n}
top_items(x, level = 5, cut = 0.3, n = 4)
```

By default items are sorted by descending `|loading|`. If your items have a
meaningful order (for example, a numbered scale you want to keep in sequence),
set `sort = FALSE`.

`top_items()` returns an object whose `$data` field is a plain data frame — a
filtered, sorted slice of `tidy(x, what = "loadings")` — so you can compute on it
if you need to:

```{r top-items-data}
ti <- top_items(x, level = 5, cut = 0.3)
head(ti$data)
```

### Cross-loadings are signal

Items that load on more than one factor are not noise to be suppressed — they
often tell you how two factors relate. Lowering the cut reveals them:

```{r top-items-cross}
top_items(x, level = 3, cut = 0.25)
```

An item that appears under two factors at the same level marks a point where the
constructs overlap. Whether that overlap is substantively meaningful or a sign of
overextraction is a judgment call — see `vignette("ackwards-suggest-k")` for the
overextraction discussion.

## The sign convention: negative does not mean "low"

Before naming anything, understand how `ackwards` orients factors. Loadings are
**sign-aligned to the primary parent** (see `?ackwards`): the level-1 factor is
anchored so its loadings sum positive, and every deeper factor is flipped so its
correlation with its primary parent is positive. This makes the diagram readable,
but it has a consequence for interpretation.

A factor's *sign* is arbitrary in the sense that flipping every loading and the
factor's orientation describes the same dimension. So a column of negative
loadings does **not** mean "low" on that construct — it means the construct's
positive pole was oriented the other way by the alignment step. Read the
*pattern* of items, not the bare sign:

```{r sign-example}
top_items(x, level = 2)
```

If a Neuroticism factor shows up with negative loadings on the anxiety items,
that is an orientation artifact of the alignment, not a "low anxiety" factor.
Name it for the construct (Neuroticism), and if the sign matters for your
downstream use, flip the scores yourself. The values shown by `top_items()` and
`tidy(what = "loadings")` are the aligned values, so they are consistent with the
diagram and the edge table.

## Naming in a hierarchy

In a flat factor solution you name each factor once. In a bass-ackwards hierarchy
you are naming factors at *every* level, and the levels are related — so the
names should be related too. The lineage tells you how.

```{r lineage}
summary(x)
```

The lineage list (`m1f1 → m2f1, m2f2 → ...`) shows each factor's primary
children. Use it to name **top-down**:

- **Upper-level factors are broader.** A level-2 factor that is the primary
  parent of two level-3 factors is the construct they share. Name it for the
  blend, not for one child. In the `bfi` data above, the level-1 factor is a
  single broad dimension; at level 2 it separates into a Neuroticism factor
  (`m2f2`, defined by the N items) and a broad factor blending the remaining four
  trait families (`m2f1`). The Big Five themselves do not appear cleanly until
  level 5 — so a substantive name for `m2f1` ("broad well-adjustment", say) is
  necessarily coarser than the names you give its descendants.

- **A split is a refinement, not a contradiction.** When a parent factor splits
  into two children, the children carve up the parent's content. Their names
  should read as specializations of the parent's name, so that reading down a
  branch tells a coherent story.

- **Watch for factors that reorganize.** A child whose strongest parent is at the
  *other* side of the level above (a crossing edge), or a factor whose primary
  edge is weak (`|r|` well below the near-1.0 values of stable dimensions), is a
  place where the structure is genuinely rearranging. The edge table makes these
  visible:

```{r edges-weak}
edges <- tidy(x, what = "edges")
primary <- edges[edges$is_primary & edges$level_to == edges$level_from + 1, ]
primary[order(abs(primary$r)), c("from", "to", "r")] |> head()
```

Edges with `|r|` near 1.0 are factors that pass through nearly unchanged — name
the child the same as the parent. The smaller `|r|` values flag where a new,
distinct construct is emerging and deserves its own name.

## Applying your names

Once you have decided on names, attach them to the diagram. `autoplot()` takes a
`node_labels` argument: a named character vector mapping factor IDs to display
strings.

Typing that vector out by hand is tedious and error-prone, so `label_template()`
generates it for you, in the same order the diagram uses, and prints an editable
`c(...)` literal you can paste straight into your script:

```{r label-template}
label_template(x)
```

Copy that literal, fill in your names, and pass it to `autoplot()`. Unspecified
IDs keep their default `m{k}f{j}` label, so you can label just the level you care
about:

```{r node-labels}
autoplot(x, node_labels = c(
  m5f1 = "Neuroticism",
  m5f2 = "Extraversion",
  m5f3 = "Openness",
  m5f4 = "Agreeableness",
  m5f5 = "Conscientiousness"
))
```

### The Forbes letter convention

Forbes (2023) labels nodes by level-letter and within-level index — `A1` for the
single level-1 factor, `B1`/`B2` at level 2, and so on. `label_template()`
produces this convention directly with `style = "forbes"`:

```{r label-forbes}
autoplot(x, node_labels = label_template(x, style = "forbes"))
```

This is useful when you want to refer to nodes by position rather than by
substantive name — for example, in a methods section that walks through the
hierarchy before interpreting it.

### Starting from a blank slate

If you would rather supply every label yourself with no defaults showing through,
`style = "blank"` gives you an all-empty scaffold to fill in:

```{r label-blank}
labs <- label_template(x, style = "blank")
labs["m5f1"] <- "N"
labs["m5f2"] <- "E"
labs["m5f3"] <- "O"
labs["m5f4"] <- "A"
labs["m5f5"] <- "C"
autoplot(x, node_labels = labs)
```

## Where to go next

This article is about *what* to put on the diagram. For *how* the diagram looks —
colours, edge thresholds, monochrome and publication styling, level labels, and
the Forbes pruned-diagram mode — see `vignette("ackwards-visualization")`.
