---
title: "Customizing the Hierarchy Diagram"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customizing the Hierarchy Diagram}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE
)
```

`autoplot.ackwards()` exposes a large number of arguments for controlling the
appearance of the hierarchy diagram. This vignette is a visual reference: each
section demonstrates one group of arguments with rendered figures so you can
see the effect before writing any code.

All options shown here are **presentation-only** — they do not change which
factors were extracted or how the between-level correlations were computed.
Options that change *which nodes appear* (`drop_pruned`, `compress_levels`) are
specific to the Forbes pruning extension and are covered in
`vignette("ackwards-forbes")`.

## Setup

```{r setup-data}
library(ackwards)

bfi <- na.omit(bfi25)
x <- ackwards(bfi, k_max = 5, cor = "polychoric")
```

The default diagram for reference:

```{r base}
autoplot(x)
```

Factors are labeled `m{k}f{j}` (level k, factor j). Two edge aesthetics carry
the between-level correlations, and each one comes with its own legend: arrow
**thickness** encodes the magnitude |r|, and edge **color** encodes the
direction (blue = positive, red–orange = negative). Level labels on the left
count factors per level.

Primary-parent edges are always positive after sign alignment; a red (negative)
edge is therefore a genuine *secondary* relationship, not an artifact.

## Encoding sign and magnitude

You choose which aesthetic carries which piece of information. `sign_by` picks
the channel for direction and `magnitude_by` picks the channel for |r|. No
aesthetic is ever mapped without a matching legend.

### `sign_by` — how direction is shown

`sign_by = "color"` (the default) uses `color_pos`/`color_neg`. `"linetype"`
draws positive edges solid and negative edges dashed, freeing color for a
single-hue figure. `"both"` uses color *and* linetype together — negatives get a
distinct double-dash so they still read in greyscale — and merges the two into a
single "Direction" legend. `"none"` drops sign encoding entirely.

```{r sign-by}
autoplot(x, sign_by = "linetype")
autoplot(x, sign_by = "both")
```

### `magnitude_by` — how |r| is shown

By default `magnitude_by = "linewidth"` maps |r| to arrow thickness with a `|r|`
legend. Set `magnitude_by = "none"` for uniform-width edges (see also
`edge_linewidth`, below, to pin a specific width).

```{r magnitude-by}
autoplot(x, magnitude_by = "none")
```

## Filtering edges

### `cut_show` — minimum |r| to display

Edges below `cut_show` are hidden entirely. Raising it produces a sparser
diagram that emphasises only the strongest connections.

```{r cut-show}
# Default cut_show = 0.3 (already shown above)
autoplot(x, cut_show = 0.5)
```

## Edge colors

### `color_pos` / `color_neg` — custom direction colors

The default blue/red palette can be replaced with any colors recognised by R.
British spellings (`colour_pos`, `colour_neg`) are accepted as aliases.

```{r colours}
autoplot(x, color_pos = "darkorchid", color_neg = "darkorange")
```

When sign is *not* encoded by color (`sign_by = "linetype"` or `"none"`), all
edges take the single `color_edge` (default black) — the basis for the Forbes
(2023) publication style (see the worked example at the end of this vignette).

## Monochrome mode

### `mono = TRUE` — a black-and-white convenience wrapper

`mono = TRUE` is shorthand for `sign_by = "linetype"` with black edges: solid
lines are positive correlations, dashed lines are negative. `magnitude_by` still
applies, so `linewidth` continues to encode |r|.

```{r mono}
autoplot(x, mono = TRUE)
```

`mono` suits black-and-white figures where the reader must distinguish positive
from negative edges. To label the edges with their exact values as well, add
`show_r = TRUE` (documented next).

## Correlation labels

### `show_r` / `r_digits` — annotate edges with r values

`show_r = TRUE` draws the rounded (signed) correlation at each edge midpoint.
`r_digits` controls the number of decimal places (default 2).

```{r show-r}
autoplot(x, show_r = TRUE)
```

```{r show-r-digits}
autoplot(x, show_r = TRUE, r_digits = 1)
```

## Node labels

### `node_labels` — rename individual factors

`node_labels` is a named character vector mapping factor IDs to display strings.
Unspecified factors keep their `m{k}f{j}` labels.

```{r node-labels}
autoplot(x, node_labels = c(
  m5f1 = "Neuro.",
  m5f2 = "Extra.",
  m5f3 = "Consc.",
  m5f4 = "Agree.",
  m5f5 = "Open."
))
```

Multi-line labels are supported via `\n`:

```{r node-labels-multiline}
autoplot(x, node_labels = c(
  m5f1 = "Neuro-\nticism",
  m5f2 = "Extra-\nversion"
))
```

### `label_template()` — generate the scaffold

Typing out every factor ID is tedious for large objects. `label_template()`
generates the full named vector in canonical diagram order and prints a
copy-pasteable `c(...)` literal you can edit and pass back to `node_labels`. It
also offers the Forbes (2023) letter convention (`"A1"`, `"B1"`, `"B2"`, …) as a
built-in style:

```{r label-template-forbes}
autoplot(x, node_labels = label_template(x, style = "forbes"))
```

For the full naming workflow — reading factors, the sign convention, and choosing
labels across the hierarchy — see `vignette("ackwards-interpret")`.

## Structural simplifications

### `primary_only = TRUE` — show only primary-parent edges

Setting `primary_only = TRUE` keeps only the single strongest edge per factor
(its primary parent), producing a clean tree. Because skip-level edges are
never primary, this also suppresses curved arcs when `pairs = "all"` was used.

```{r primary-only}
autoplot(x, primary_only = TRUE)
```

## Level labels

### `show_level_labels` / `level_label_size`

Level labels ("1 factor", "2 factors", …) are shown by default on the left
margin. They can be hidden or resized.

```{r level-labels-off}
autoplot(x, show_level_labels = FALSE)
```

```{r level-labels-size}
autoplot(x, show_level_labels = TRUE, level_label_size = 4)
```

## Arrowheads

### `show_arrows = FALSE` — plain line ends

By default, edges end with closed arrowheads. Setting `show_arrows = FALSE`
draws plain line ends. This applies to both straight edges and curved skip-level
arcs.

```{r no-arrows}
autoplot(x, show_arrows = FALSE)
```

## Edge width

### `edge_linewidth` — uniform vs. |r|-scaled width

By default, edge width is proportional to |r| (via `magnitude_by`). A numeric
`edge_linewidth` draws every edge at that constant width and removes the `|r|`
legend — like `magnitude_by = "none"`, but at a width you choose.

```{r edge-linewidth}
autoplot(x, edge_linewidth = 0.7)
```

## Layout orientation

### `direction = "horizontal"` — left-to-right layout

By default levels stack top-to-bottom (level 1 at top). `direction =
"horizontal"` lays them out left-to-right (level 1 at left), which fits wide
slides and posters; the level labels move to the bottom margin.

```{r direction, fig.width=8, fig.height=4}
autoplot(x, direction = "horizontal")
```

## Legend

### `legend = FALSE` — suppress all guides

`legend = FALSE` removes all legends from the plot. Most useful when the diagram
is self-explanatory or the legend duplicates information conveyed by labels.

```{r no-legend}
autoplot(x, legend = FALSE)
```

## Worked example: publication-ready figure

The following call reproduces the visual style of Forbes (2023): black lines of
uniform weight, plain line ends, correlation labels, and no legend. Setting both
direction colors to black yields a single-hue figure, and `legend = FALSE`
suppresses the now-redundant key.

```{r pub-figure}
autoplot(x,
  color_pos      = "black",
  color_neg      = "black",
  edge_linewidth = 0.6,
  show_arrows    = FALSE,
  show_r         = TRUE,
  legend         = FALSE
)
```

Combining with `node_labels` names the factors for the final figure:

```{r pub-figure-labeled}
autoplot(x,
  color_pos = "black",
  color_neg = "black",
  edge_linewidth = 0.6,
  show_arrows = FALSE,
  show_r = TRUE,
  legend = FALSE,
  node_labels = c(
    m5f1 = "Neuro.",
    m5f2 = "Extra.",
    m5f3 = "Consc.",
    m5f4 = "Agree.",
    m5f5 = "Open."
  )
)
```

For the pruned-factor variant of this figure (nodes omitted, spanning arrows)
see `vignette("ackwards-forbes")`.

## Saving plots

`autoplot()` returns an ordinary `ggplot` object, so save it with
`ggplot2::ggsave()`:

```{r ggsave, eval=FALSE}
p <- autoplot(x, direction = "horizontal")
ggplot2::ggsave("hierarchy.png", p, width = 9, height = 5, dpi = 300)
```

`ackwards` does not re-export `ggsave()` — that would move `ggplot2` from
Suggests into Imports — so call it from `ggplot2` directly.

---

## Diagnostic scree / criteria plot: `autoplot.suggest_k()`

`suggest_k()` also has its own `autoplot()` method, producing a multi-panel
scree / parallel-analysis / VSS diagnostic. It is documented in depth in
`vignette("ackwards-suggest-k")`; here we only note that the same `autoplot()`
generic covers it.

```{r suggest_k_plot, fig.height=8}
sk <- suggest_k(bfi, seed = 42)
autoplot(sk)
```

## References

Forbes, M. K. (2023). Improving hierarchical models of individual differences:
An extension of Goldberg's bass-ackward method. *Psychological Methods*.
https://doi.org/10.1037/met0000546
