---
title: "Choosing an Engine: PCA, EFA, and ESEM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing an Engine: PCA, EFA, and ESEM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
has_lavaan <- requireNamespace("lavaan", quietly = TRUE)
```

**ackwards** supports three factor extraction engines. They share the same
downstream machinery — the same rotation, the same tenBerge scoring weights, the
same between-level correlation algebra — but differ in their statistical model
and what they report. This vignette explains when each one is appropriate and
what the differences look like in practice.

## The three engines at a glance

| | `"pca"` | `"efa"` | `"esem"` |
|-|---------|---------|----------|
| **What it models** | Total item variance | Common (latent) variance | Common variance via lavaan |
| **Communalities** | All 1.0 (by definition) | Estimated from data | Estimated from data |
| **Fit indices** | Eigenvalues only | chi, RMSEA, TLI, BIC | CFI, TLI, RMSEA, SRMR + chi |
| **Loading SEs** | No | No | Yes (WLSMV for ordinal) |
| **Speed** | Fast | Moderate | Slowest |
| **Best for** | Exploration, large k | Latent-factor inference | Model evaluation, ordinal data |

All three produce the same labels (`m{k}f{j}`), the same `tidy()` / `glance()`
/ `augment()` interface, and comparable between-level edges for well-structured
data. The hierarchy they reveal is usually the same; the statistical guarantees
differ.

## Setup

```{r setup-data}
library(ackwards)
bfi <- na.omit(bfi25)
```

We use the BFI-25 with polychoric correlations throughout so that differences in
output reflect the engine, not the correlation basis.

## PCA: components from total variance

PCA extracts **principal components** — linear combinations of the observed
variables that capture maximum variance, including measurement error. Every item
is modeled with communality 1.0: the components account for 100% of each item's
variance. This is not a true latent variable model; it is a data reduction
method.

In the bass-ackwards context, PCA is the natural default. It is fast, always
converges, and produces eigenvalues that can guide the choice of k. Waller
(2007) showed that the between-level algebra (`W'RW`) holds exactly for
components, making the edges algebraically exact rather than approximated from
materialized scores.

```{r pca}
x_pca <- ackwards(bfi, k_max = 3, cor = "polychoric")
x_pca
```

The "fit" for PCA is just the eigenvalue of each component — the amount of
variance it captures. There are no chi-square tests, no RMSEA, no model
rejection.

```{r pca-fit}
tidy(x_pca, what = "fit")
```

## EFA: factors from common variance

EFA extracts **latent factors** that model only the variance shared among items.
Each item retains a unique variance (communality < 1.0) that the factors do not
explain. This is the classical common-factor model, and it is more appropriate
than PCA when you believe the items are fallible indicators of latent constructs
rather than the constructs themselves.

```{r efa}
x_efa <- ackwards(bfi, k_max = 3, engine = "efa", cor = "polychoric")
x_efa
```

EFA produces genuine goodness-of-fit indices. These tell you whether the k
factors are sufficient to reproduce the observed correlation matrix within
sampling error.

```{r efa-fit}
tidy(x_efa, what = "fit")
```

The RMSEA values here are large (> 0.10), indicating that 1–3 factors do not
fully account for the BFI item correlations — unsurprising, because the true
structure is 5 factors. Fit improves steadily from k = 1 to k = 3, which is
exactly the kind of evidence bass-ackwards analysis is designed to make visible.

### How close are EFA and PCA loadings?

For clean, continuous data with moderate-to-strong factor structure, EFA and
PCA loadings are highly correlated but not identical. EFA loadings are
systematically somewhat smaller because they model only the common variance;
PCA inflates loadings by fitting noise alongside signal.

The table below pairs the `k = 3` loadings from each engine — built by stacking
`tidy(x_pca, what = "loadings")` and the EFA equivalent — for six representative
items (two each from the Neuroticism, Extraversion, and Conscientiousness
families).

```{r loadings-compare, echo = FALSE}
anchors <- c("N1", "N2", "E1", "E2", "C1", "C2")
loadings <- rbind(
  cbind(engine = "pca", tidy(x_pca, what = "loadings")),
  cbind(engine = "efa", tidy(x_efa, what = "loadings"))
)
loadings <- loadings[loadings$level == 3 & loadings$item %in% anchors, ]
loadings <- loadings[order(loadings$factor, loadings$item), ]
knitr::kable(
  loadings[, c("engine", "factor", "item", "loading")],
  digits = 2, row.names = FALSE,
  caption = "PCA vs EFA loadings for anchor items (k = 3)"
)
```

EFA loadings for the same items are consistently a few points lower — the PCA
loadings include some noise variance that EFA partitions into uniqueness. The
factor structure (which items define which factor) is unchanged.

## ESEM: EFA with full model diagnostics

ESEM (exploratory structural equation modeling, Asparouhov & Muthén, 2009) fits
the same common-factor model as EFA but uses **lavaan** as the engine. This
unlocks two capabilities that EFA cannot provide:

1. **Standard errors for every loading**, enabling confidence intervals and
   significance tests.
2. **The WLSMV estimator** for ordinal data, which is the appropriate
   maximum-likelihood-adjacent estimator for categorical indicators. When
   `cor = "polychoric"` is set with `engine = "esem"`, WLSMV is used
   automatically.

```{r esem-check, echo = FALSE, results = "asis"}
if (!has_lavaan) {
  cat(
    "> **Note:** The ESEM examples below require the **lavaan** package,",
    "which is not currently installed. Install it with",
    "`install.packages(\"lavaan\")` to run these chunks.\n"
  )
}
```

```{r esem, eval = has_lavaan}
x_esem <- ackwards(bfi, k_max = 3, engine = "esem", cor = "polychoric")
x_esem
```

ESEM fit indices include CFI and SRMR in addition to RMSEA and TLI, giving a
richer picture of model adequacy.

```{r esem-fit, eval = has_lavaan}
tidy(x_esem, what = "fit")
```

### Loading standard errors

The unique output from ESEM is the rotation-aware **standard error** of every
loading. `tidy(what = "loadings_se")` returns them in the same long format as
`tidy(what = "loadings")`, with an `se` column in place of `loading`:

```{r esem-se, eval = has_lavaan}
se <- tidy(x_esem, what = "loadings_se")
head(se)
```

These SEs allow you to construct confidence intervals: loading ± 1.96 × SE
gives an approximate 95% CI. For the BFI with > 2,000 participants the SEs are
small; with smaller samples they become important for judging which loadings are
meaningfully non-zero. (`tidy(what = "loadings_se")` errors for PCA and EFA
objects, which carry no loading standard errors.)

## How much do the edges differ?

The primary output of bass-ackwards analysis is the between-level edges. For
well-structured, continuous data, all three engines should agree closely on the
hierarchy.

The table stacks each engine's primary-parent edges — `tidy(what = "edges",
primary_only = TRUE)` — at every level transition:

```{r edges-compare, echo = FALSE}
edges <- rbind(
  cbind(engine = "pca", tidy(x_pca, what = "edges", primary_only = TRUE)),
  cbind(engine = "efa", tidy(x_efa, what = "edges", primary_only = TRUE))
)
edges <- edges[order(edges$to), c("engine", "from", "to", "r")]
knitr::kable(
  edges,
  digits = 2, row.names = FALSE,
  caption = "Primary-parent edges: PCA vs EFA"
)
```

The r values are very close between engines: the hierarchy that PCA reveals is
essentially the same hierarchy that EFA reveals. This convergence across methods
is reassuring — it suggests the structure is real and not an artifact of the
extraction method.

When the engines disagree on edges, that is itself informative: it usually
indicates factors whose definition depends on whether you account for
measurement error (EFA/ESEM) or not (PCA).

## Choosing an engine

| Situation | Recommendation |
|-----------|---------------|
| Exploratory, large k, unknown structure | Start with `"pca"` |
| Latent-variable theory, want to test model fit | `"efa"` |
| Ordinal items + model fit + loading SEs | `"esem"` with `cor = "polychoric"` |
| Replicating Goldberg (2006) or `psych::bassAckward()` | `"pca"`, `fm = "pca"` |
| Publication with formal model evaluation | `"esem"` |

A practical workflow: start with PCA to get a feel for the hierarchy and choose
k. Switch to EFA or ESEM to confirm and report. If PCA and EFA/ESEM edges
agree, you have robust evidence for the hierarchy; if they disagree, investigate
why.

## Missing data

`ackwards()` accepts a `missing` argument with three options.

**`"pairwise"` (default).** Use all available observations pairwise. The exact
behaviour depends on the engine and estimator:

- **PCA/EFA:** `stats::cor(use = "pairwise.complete.obs")` — uses all rows that
  contribute to each variable pair. MCAR-valid; N = total rows.
- **ESEM WLSMV/ULSMV (ordinal):** lavaan `missing = "available.cases"` —
  computes polychoric thresholds and correlations from every row contributing to
  each pair. MCAR-valid; uses the full N, not just complete cases. This is the
  honest interpretation of "pairwise" for ordinal data.
- **ESEM ML/MLR (continuous):** lavaan uses listwise deletion internally while
  edge correlations are computed from a separately-computed pairwise correlation
  matrix. This minor inconsistency (fit statistics at complete-case N, edges at
  full N) is documented in `$meta`. Prefer `"listwise"` or `"fiml"` when
  missingness is substantial.

A warning is emitted whenever incomplete rows are detected with this option.

```r
# Default: pairwise (warns if NAs present)
x <- ackwards(data_with_nas, k_max = 4)
```

**`"listwise"`.** Data are reduced to complete cases before *all* downstream
steps — correlation matrix, engine fitting, and edges — so the three quantities
are fully consistent. `x$n_obs` and `x$meta$n_complete` both reflect the
reduced sample size. Valid for all three engines.

```r
# Consistent complete-case analysis
x <- ackwards(data_with_nas, k_max = 4, missing = "listwise")
x$n_obs         # complete-case N
x$meta$missing  # "listwise"
```

**`"fiml"`.** Full Information Maximum Likelihood — available only for
`engine = "esem"` with `estimator = "ML"` or `"MLR"`. FIML uses information
from all rows (including those with partial data) when estimating loadings and
fit. Edge correlations are derived from lavaan's FIML-estimated saturated model
(the h1 unrestricted model), ensuring that fits and edges use the same
information. Note: FIML improves *estimation* but does not impute item
responses; score materialisation (`keep_scores = TRUE`) still yields `NA` for
incomplete rows.

```r
# FIML for ESEM with continuous data
x <- ackwards(data_with_nas, k_max = 4, engine = "esem",
              estimator = "ML", missing = "fiml")
```

`"fiml"` errors clearly for unsupported combinations:

```r
# Errors: PCA and EFA are correlation-based, not raw-data likelihood
ackwards(data, k_max = 4, engine = "pca", missing = "fiml")

# Errors: WLSMV is limited-information WLS, no FIML extension
ackwards(data, k_max = 4, engine = "esem", cor = "polychoric",
         missing = "fiml")
```

### Which option to use?

| Situation | Recommendation |
|-----------|----------------|
| Continuous data, little missingness | `"pairwise"` (default) |
| Ordinal data + WLSMV, any missingness | `"pairwise"` (uses `available.cases` — MCAR-valid, full N) |
| Want consistent fit statistics and edges (continuous ML/MLR) | `"listwise"` |
| ESEM ML/MLR, meaningful missingness, want all rows used in estimation | `"fiml"` |
| MAR-valid with ordinal (not yet built-in) | MI via `lavaan.mi` or `mirt` |

## Correlation-matrix input

When you have a pre-computed correlation matrix — a published table, a
polychoric matrix computed externally, or a subset you want to analyse without
refitting — you can pass it directly to `ackwards()` or `suggest_k()`. The
matrix is auto-detected from its shape (square, symmetric, unit diagonal).

```{r cor-matrix-input, eval = FALSE}
R <- cor(bfi25, use = "pairwise.complete.obs")

# PCA from a correlation matrix (n_obs optional for PCA; required for EFA)
x_R <- ackwards(R, k_max = 5)
# EFA requires n_obs for fit indices:
x_efa_R <- ackwards(R, k_max = 5, engine = "efa", n_obs = 875L)

# Edges are identical to the raw-data run (same W'RW algebra):
x_d <- ackwards(bfi25, k_max = 5)
all.equal(tidy(x_R)$r, tidy(x_d)$r) # TRUE within floating-point tolerance
```

### Constraints

| Constraint | Detail |
|---|---|
| **Engine** | `"pca"` and `"efa"` only — `"esem"` errors (lavaan needs raw data) |
| **`n_obs`** | Required for `"efa"`; optional for `"pca"` (stored as `NA`) |
| **`cor` argument** | Ignored (basis is fixed); warns if set explicitly |
| **`missing` argument** | Ignored; warns if set explicitly |
| **Factor scores** | `keep_scores = TRUE`, `augment()`, `tidy(what = "scores")` all error |
| **`$cor` field** | Stored as `NA`; shown as `"(user-supplied matrix)"` in print |

### suggest_k() with a correlation matrix

```{r cor-matrix-suggest, eval = FALSE}
sk_R <- suggest_k(R, n_obs = 875L)
# CD is skipped (resampling requires raw item distributions)
# PA, MAP, and VSS run normally
```

## References

Goldberg, L. R. (2006). Doing it all bass-ackwards. *Journal of Research in
Personality*, *40*(4), 347–358.

Waller, N. G. (2007). A general method for computing hierarchical component
structures by Bass-Ackward factor analysis. *Journal of Research in
Personality*, *41*(4), 745–752.

Asparouhov, T., & Muthén, B. (2009). Exploratory structural equation modeling.
*Structural Equation Modeling*, *16*(3), 397–438.
