---
title: "Choosing an Engine: PCA, EFA, and ESEM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing an Engine: PCA, EFA, and ESEM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
has_lavaan <- requireNamespace("lavaan", quietly = TRUE)
has_gt <- requireNamespace("gt", quietly = TRUE)
```

**ackwards** supports three factor extraction engines. They share the same
downstream machinery — the same rotation, the same tenBerge scoring weights, the
same between-level correlation algebra — but differ in their statistical model
and what they report. This vignette explains when each one is appropriate and
what the differences look like in practice.

## The three engines at a glance

| | `"pca"` | `"efa"` | `"esem"` |
|-|---------|---------|----------|
| **What it models** | Total item variance | Common (latent) variance | Common (latent) variance |
| **Engine substrate** | `psych::principal()` | `psych::fa()` | `lavaan` |
| **Communalities** | All 1.0 (by definition) | Estimated from data | Estimated from data |
| **Correlations** | Pearson or polychoric | Pearson or polychoric | Pearson or polychoric |
| **Estimators** | Eigen-decomposition | OLS / minres / ML (via `psych`) | ML, MLR — continuous (+ FIML for missing data); WLSMV, ULSMV — ordinal |
| **Fit indices** | Eigenvalues only | χ², RMSEA, TLI, BIC | CFI, TLI, RMSEA, SRMR, χ² |
| **Loading SEs** | No | No | Yes |
| **Speed** | Fast | Moderate | Slowest |
| **Best for** | Exploration, large k | Latent-factor inference | Model evaluation, loading SEs, ordinal *or* continuous |

All three produce the same labels (`m{k}f{j}`), the same `tidy()` / `glance()`
/ `augment()` interface, and comparable between-level edges for well-structured
data. The hierarchy they reveal is usually the same; the statistical guarantees
differ.

## Setup

```{r setup-data}
library(ackwards)
bfi <- na.omit(bfi25)
```

We use the BFI-25 with polychoric correlations throughout so that differences in
output reflect the engine, not the correlation basis.

## PCA: components from total variance

PCA extracts **principal components** — linear combinations of the observed
variables that capture maximum variance, including measurement error. Every item
is modeled with communality 1.0: the components account for 100% of each item's
variance. This is not a true latent variable model; it is a data reduction
method.

In the bass-ackwards context, PCA is the natural default. It is fast, always
converges, and produces eigenvalues that can guide the choice of k. Waller
(2007) showed that the between-level algebra (`W'RW`) holds exactly for
components, making the edges algebraically exact rather than approximated from
materialized scores.

```{r pca}
x_pca <- ackwards(bfi, k_max = 3, cor = "polychoric")
x_pca
```

The "fit" for PCA is just the eigenvalue of each component — the amount of
variance it captures. There are no chi-square tests, no RMSEA, no model
rejection.

```{r pca-fit}
tidy(x_pca, what = "fit")
```

## EFA: factors from common variance

EFA extracts **latent factors** that model only the variance shared among items.
Each item retains a unique variance (communality < 1.0) that the factors do not
explain. This is the classical common-factor model, and it is more appropriate
than PCA when you believe the items are fallible indicators of latent constructs
rather than the constructs themselves.

```{r efa}
x_efa <- ackwards(bfi, k_max = 3, engine = "efa", cor = "polychoric")
x_efa
```

EFA produces genuine goodness-of-fit indices. These tell you whether the k
factors are sufficient to reproduce the observed correlation matrix within
sampling error.

```{r efa-fit}
tidy(x_efa, what = "fit")
```

The RMSEA values here are large (> 0.10), indicating that 1–3 factors do not
fully account for the BFI item correlations — unsurprising, because the true
structure is 5 factors. Fit improves steadily from k = 1 to k = 3, which is
exactly the kind of evidence bass-ackwards analysis is designed to make visible.

### How close are EFA and PCA loadings?

For clean, continuous data with moderate-to-strong factor structure, EFA and
PCA loadings are highly correlated but not identical. EFA loadings are
systematically somewhat smaller because they model only the common variance;
PCA inflates loadings by fitting noise alongside signal.

The table below compares primary loadings — the loading of each item on its
dominant factor — for six representative items (two each from the Neuroticism,
Extraversion, and Conscientiousness families) at k = 3. The Δ column is the
teaching point: how much smaller EFA loadings are in absolute value once
measurement error is partitioned into uniqueness. Using |EFA| − |PCA| keeps
the attenuation consistently negative regardless of loading sign.

```{r loadings-compare, echo = FALSE}
# Primary loading = loading on the factor with the highest |loading| for that item
anchors <- c("N1", "N2", "E1", "E2", "C1", "C2")

primary_loading <- function(x, level_k) {
  d <- tidy(x, what = "loadings")
  d <- d[d$level == level_k & d$item %in% anchors, ]
  do.call(rbind, lapply(split(d, d$item), function(r) r[which.max(abs(r$loading)), ]))
}

pca_p <- primary_loading(x_pca, 3)
efa_p <- primary_loading(x_efa, 3)

# Verify factor and sign correspondence before differencing
stopifnot(
  all(pca_p$factor[order(pca_p$item)] == efa_p$factor[order(efa_p$item)]),
  all(sign(pca_p$loading[order(pca_p$item)]) == sign(efa_p$loading[order(efa_p$item)]))
)

wide <- merge(
  pca_p[, c("factor", "item", "loading")],
  efa_p[, c("item", "loading")],
  by = "item", suffixes = c("_pca", "_efa")
)
wide$delta <- abs(wide$loading_efa) - abs(wide$loading_pca)
wide <- wide[order(wide$factor, wide$item), c("item", "factor", "loading_pca", "loading_efa", "delta")]

if (has_gt) {
  gt::gt(wide) |>
    gt::tab_header(title = "PCA vs EFA: primary loadings for anchor items (k = 3)") |>
    gt::tab_spanner(label = "Loading", columns = c(loading_pca, loading_efa, delta)) |>
    gt::cols_label(
      item = "Item",
      factor = "Factor",
      loading_pca = "PCA",
      loading_efa = "EFA",
      delta = "Δ (|EFA| − |PCA|)"
    ) |>
    gt::fmt_number(columns = c(loading_pca, loading_efa, delta), decimals = 2) |>
    gt::tab_style(
      style     = gt::cell_text(weight = "bold"),
      locations = gt::cells_column_labels(columns = delta)
    ) |>
    gt::tab_footnote(
      footnote = paste0(
        "Factor assignment and sign verified to match between engines.",
        " Delta uses |EFA| - |PCA| so attenuation is always negative regardless of loading sign."
      ),
      locations = gt::cells_column_labels(columns = delta)
    )
} else {
  knitr::kable(
    wide,
    digits = 2, row.names = FALSE,
    col.names = c("Item", "Factor", "PCA", "EFA", "Δ (|EFA| − |PCA|)"),
    caption = "PCA vs EFA: primary loadings for anchor items (k = 3)"
  )
}
```

EFA loadings for the same items are consistently a few points lower — the PCA
loadings include some noise variance that EFA partitions into uniqueness. The
factor structure (which items define which factor) is unchanged.

## ESEM: EFA with full model diagnostics

ESEM (exploratory structural equation modeling, Asparouhov & Muthén, 2009) fits
the same common-factor model as EFA but uses **lavaan** as the engine. ESEM is
not an ordinal-only tool: it handles **continuous** items with maximum-likelihood
estimators (ML, MLR — the default for continuous data) and **ordinal** items with
WLSMV, and it unlocks three capabilities that EFA cannot provide:

1. **Standard errors for every loading**, enabling confidence intervals and
   significance tests — for continuous *and* ordinal data.
2. **Full maximum-likelihood estimation** for continuous data (ML/MLR), including
   **FIML** for missing data (`missing = "fiml"`), which uses all partially
   observed rows rather than deleting them.
3. **The WLSMV estimator** for ordinal data, the appropriate
   maximum-likelihood-adjacent estimator for categorical indicators. When
   `cor = "polychoric"` is set with `engine = "esem"`, WLSMV is used
   automatically.

```{r esem-check, echo = FALSE, results = "asis"}
if (!has_lavaan) {
  cat(
    "> **Note:** The ESEM examples below require the **lavaan** package,",
    "which is not currently installed. Install it with",
    "`install.packages(\"lavaan\")` to run these chunks.\n"
  )
}
```

```{r esem, eval = has_lavaan}
x_esem <- ackwards(bfi, k_max = 3, engine = "esem", cor = "polychoric")
x_esem
```

ESEM fit indices include CFI and SRMR in addition to RMSEA and TLI, giving a
richer picture of model adequacy. See the "Per-level fit" section below for how
to report and interpret these indices.

### Loading standard errors and confidence intervals

The unique output from ESEM is the rotation-aware **standard error** of every
loading. These SEs are now returned as part of `tidy(what = "loadings")`,
alongside `ci_lower` and `ci_upper` columns:

```{r esem-se, eval = has_lavaan}
ld <- tidy(x_esem, what = "loadings")
head(ld)
```

The intervals are computed as loading ± *z* × SE (default 95%; set
`conf_level = 0.99` for wider intervals). For the BFI with > 2,000 participants
the SEs are small; with smaller samples they are important for judging which
loadings are meaningfully non-zero. For PCA and EFA objects the `se`,
`ci_lower`, and `ci_upper` columns are present but `NA` — those engines carry no
loading SEs.

## Per-level fit: what it tells you (and what it doesn't)

### The key distinction

Bass-ackwards produces a **series of independent factor solutions**, not a
fitted hierarchical model. The between-level edges are descriptive correlations
between factor scores — they have no sampling distribution of their own. Per-level
fit indices therefore describe something narrower: **does a k-factor model
adequately reproduce the items at this level?**

That is a real, bounded question. A level that fits terribly is one you
shouldn't over-interpret — the k factors are not cleanly separating the items.
A level that fits well tells you the factor structure at that depth is stable. But
good fit at k = 3 does **not** validate the edges connecting k = 3 to k = 2; it
only says the k = 3 solution itself is trustworthy. Keep that boundary in mind
whenever you report or interpret fit.

The converse question comes up often: *if the k = 3 solution fits badly, should I
distrust the edges connecting k = 2 to k = 3?* The honest answer is nuanced. The
edge correlation is still a **faithful description** of the relationship between
the k = 2 and k = 3 factor scores as extracted — the arithmetic is not wrong. What
poor fit undermines is the **interpretation** of the k = 3 factors themselves: if
three factors don't cleanly reproduce the items, then "factor `m3f2`" is a shakier
construct, and any edge *incident to it* inherits that shakiness. So a badly-fitting
level does weaken the edges touching it — not because the correlation is
miscomputed, but because one of the things it connects is poorly defined. The edges
between two *well*-fitting levels are on firmer ground than edges touching a
poorly-fitting one.

### Reporting fit with `tidy()` and `autoplot()`

`tidy(what = "fit")` returns the raw long table. For reporting, `format = "wide"`
gives one row per level:

```{r fit-wide, eval = has_lavaan}
tidy(x_esem, what = "fit", format = "wide")
```

`tidy()` does not flag rows against a threshold — the Hu & Bentler (1999)
conventional cutoffs (CFI/TLI ≥ .95, RMSEA ≤ .06, SRMR ≤ .08) are conventional
and contested, so a pass/fail column would overstate their authority. Instead
they appear only as visual/inline reference points in `autoplot()` and
`summary()`:

> **Thresholds are conventional and contested.** They were derived from specific
> simulation conditions (continuous, well-distributed items; balanced designs).
> WLSMV fit for ordinal data tends to produce lower CFI and higher RMSEA than ML
> on the same underlying structure; do not interpret WLSMV cutoffs as strictly as
> ML-based rules. Use them as a rough orientation, not a gatekeeping criterion.

`autoplot(x, what = "fit")` visualises the trajectory across levels, with cutoff
reference lines:

```{r fit-plot, eval = has_lavaan, fig.width = 6, fig.height = 5}
autoplot(x_esem, what = "fit")
```

The shape of the trajectory matters as much as the absolute values: a sharp
improvement from k = 2 to k = 3 suggests the third factor is capturing genuine
signal; flat or worsening indices suggest adding another level is splitting noise.

The examples in this vignette deliberately stop at `k_max = 3` so the ESEM chunks
build quickly and the tables stay legible — not because the BFI hierarchy ends
there. For these data `suggest_k()` points to roughly **k = 5** (see the
`vignette("ackwards-suggest-k")`), so a real analysis would extend the trajectory
further and read the fit curve across all five levels. The truncated plot here
shows the *mechanics* of reading a fit trajectory, not the recommended depth for
the BFI.

`glance()` now also carries the deepest-level fit for quick inspection:

```{r glance-fit, eval = has_lavaan}
glance(x_esem)
```

### Should you care about fit in a bass-ackwards workflow?

It depends on your goal:

- **Exploratory** (finding the hierarchical structure): fit is a secondary check.
  Start with PCA for speed, confirm with EFA or ESEM. If a level's fit is poor,
  consider whether you have too many factors at that depth, or whether the items
  at that level are genuinely multidimensional.
- **Confirmatory / publication**: fit is table-stakes for ESEM or EFA. Report
  per-level CFI, TLI, RMSEA (and SRMR for ESEM) alongside the hierarchy. The
  wide table and `autoplot(what = "fit")` are designed for this.
- **Ordinal data**: WLSMV (ESEM) gives fit indices appropriate for categorical
  items; EFA's RMSEA/TLI under Pearson correlation is a rougher diagnostic.

The bottom line: **per-level fit qualifies each level of the hierarchy; it does
not bless the hierarchy as a whole.** Use it to decide how deep the structure is
credibly resolved, not to claim the overall model is "good".

## How much do the edges differ?

The primary output of bass-ackwards analysis is the between-level edges. For
well-structured, continuous data, all three engines should agree closely on the
hierarchy.

The table below compares the primary-parent edge strength for every adjacent
level transition. The Δ column is the shift in connection strength
(|EFA| − |PCA|) — a direct, sign-robust measure of how much the latent-variable
model changes your inference about the hierarchy.

```{r edges-compare, echo = FALSE}
pca_e <- tidy(x_pca, what = "edges", primary_only = TRUE)
efa_e <- tidy(x_efa, what = "edges", primary_only = TRUE)

wide_e <- merge(
  pca_e[, c("from", "to", "r")],
  efa_e[, c("from", "to", "r")],
  by = c("from", "to"), all = TRUE, suffixes = c("_pca", "_efa")
)
wide_e$delta <- abs(wide_e$r_efa) - abs(wide_e$r_pca)
wide_e <- wide_e[order(wide_e$to), c("from", "to", "r_pca", "r_efa", "delta")]

if (has_gt) {
  gt::gt(wide_e) |>
    gt::tab_header(title = "Primary-parent edges: PCA vs EFA") |>
    gt::tab_spanner(label = "Edge strength (r)", columns = c(r_pca, r_efa, delta)) |>
    gt::cols_label(
      from  = "From",
      to    = "To",
      r_pca = "PCA",
      r_efa = "EFA",
      delta = "Δ (EFA − PCA)"
    ) |>
    gt::fmt_number(columns = c(r_pca, r_efa, delta), decimals = 2) |>
    gt::tab_style(
      style     = gt::cell_text(weight = "bold"),
      locations = gt::cells_column_labels(columns = delta)
    ) |>
    gt::tab_footnote(
      footnote  = "NA in either column means the engines disagree on the primary parent for that factor.",
      locations = gt::cells_column_labels(columns = delta)
    )
} else {
  knitr::kable(
    wide_e,
    digits = 2, row.names = FALSE,
    col.names = c("From", "To", "PCA", "EFA", "Δ (|EFA| − |PCA|)"),
    caption = "Primary-parent edges: PCA vs EFA"
  )
}
```

The r values are very close between engines: the hierarchy that PCA reveals is
essentially the same hierarchy that EFA reveals. This convergence across methods
is reassuring — it suggests the structure is real and not an artifact of the
extraction method.

How clean this convergence looks depends on how strong the underlying structure
is. The simulated `sim16` dataset (`?sim16`) is an *idealized* case — its planted
1 → 2 → 4 hierarchy is strong enough that engines and `suggest_k()` criteria agree
almost perfectly. Real data is messier: for `bfi25` the `suggest_k()` criteria
span k = 4–6 even though the engines agree on the edges. Treat clean cross-method
consensus as the best case, not the norm; the `vignette("ackwards-suggest-k")`
develops this idealized-vs-realistic contrast in full.

When the engines disagree on edges, that is itself informative: it usually
indicates factors whose definition depends on whether you account for
measurement error (EFA/ESEM) or not (PCA).

## Choosing an engine

| Situation | Recommendation |
|-----------|---------------|
| Exploratory, large k, unknown structure | Start with `"pca"` |
| Latent-variable theory, want to test model fit | `"efa"` |
| Ordinal items + model fit + loading SEs | `"esem"` with `cor = "polychoric"` |
| Replicating Goldberg (2006) or `psych::bassAckward()` | `"pca"`, `fm = "pca"` |
| Publication with formal model evaluation | `"esem"` |

A practical workflow: start with PCA to get a feel for the hierarchy and choose
k. Switch to EFA or ESEM to confirm and report. If PCA and EFA/ESEM edges
agree, you have robust evidence for the hierarchy; if they disagree, investigate
why.

## Missing data

`ackwards()` accepts a `missing` argument with three options.

**`"pairwise"` (default).** Use all available observations pairwise. The exact
behaviour depends on the engine and estimator:

- **PCA/EFA:** `stats::cor(use = "pairwise.complete.obs")` — uses all rows that
  contribute to each variable pair. MCAR-valid; N = total rows.
- **ESEM WLSMV/ULSMV (ordinal):** lavaan `missing = "available.cases"` —
  computes polychoric thresholds and correlations from every row contributing to
  each pair. MCAR-valid; uses the full N, not just complete cases. This is the
  honest interpretation of "pairwise" for ordinal data.
- **ESEM ML/MLR (continuous):** lavaan uses listwise deletion internally while
  edge correlations are computed from a separately-computed pairwise correlation
  matrix. This minor inconsistency (fit statistics at complete-case N, edges at
  full N) is documented in `$meta`. Prefer `"listwise"` or `"fiml"` when
  missingness is substantial.

A warning is emitted whenever incomplete rows are detected with this option.

```r
# Default: pairwise (warns if NAs present)
x <- ackwards(data_with_nas, k_max = 4)
```

**`"listwise"`.** Data are reduced to complete cases before *all* downstream
steps — correlation matrix, engine fitting, and edges — so the three quantities
are fully consistent. `x$n_obs` and `x$meta$n_complete` both reflect the
reduced sample size. Valid for all three engines.

```r
# Consistent complete-case analysis
x <- ackwards(data_with_nas, k_max = 4, missing = "listwise")
x$n_obs         # complete-case N
x$meta$missing  # "listwise"
```

**`"fiml"`.** Full Information Maximum Likelihood — available only for
`engine = "esem"` with `estimator = "ML"` or `"MLR"`. FIML uses information
from all rows (including those with partial data) when estimating loadings and
fit. Edge correlations are derived from lavaan's FIML-estimated saturated model
(the h1 unrestricted model), ensuring that fits and edges use the same
information. Note: FIML improves *estimation* but does not impute item
responses; score materialisation (`keep_scores = TRUE`) still yields `NA` for
incomplete rows.

```r
# FIML for ESEM with continuous data
x <- ackwards(data_with_nas, k_max = 4, engine = "esem",
              estimator = "ML", missing = "fiml")
```

`"fiml"` errors clearly for unsupported combinations:

```r
# Errors: PCA and EFA are correlation-based, not raw-data likelihood
ackwards(data, k_max = 4, engine = "pca", missing = "fiml")

# Errors: WLSMV is limited-information WLS, no FIML extension
ackwards(data, k_max = 4, engine = "esem", cor = "polychoric",
         missing = "fiml")
```

### Which option to use?

| Situation | Recommendation |
|-----------|----------------|
| Continuous data, little missingness | `"pairwise"` (default) |
| Ordinal data + WLSMV, any missingness | `"pairwise"` (uses `available.cases` — MCAR-valid, full N) |
| Want consistent fit statistics and edges (continuous ML/MLR) | `"listwise"` |
| ESEM ML/MLR, meaningful missingness, want all rows used in estimation | `"fiml"` |
| MAR-valid with ordinal (not yet built-in) | MI via `lavaan.mi` or `mirt` |

## Correlation-matrix input

When you have a pre-computed correlation matrix — a published table, a
polychoric matrix computed externally, or a subset you want to analyse without
refitting — you can pass it directly to `ackwards()` or `suggest_k()`. The
matrix is auto-detected from its shape (square, symmetric, unit diagonal).

```{r cor-matrix-input, eval = FALSE}
R <- cor(bfi25, use = "pairwise.complete.obs")

# PCA from a correlation matrix (n_obs optional for PCA; required for EFA)
x_R <- ackwards(R, k_max = 5)
# EFA requires n_obs for fit indices:
x_efa_R <- ackwards(R, k_max = 5, engine = "efa", n_obs = 875L)

# Edges are identical to the raw-data run (same W'RW algebra):
x_d <- ackwards(bfi25, k_max = 5)
all.equal(tidy(x_R)$r, tidy(x_d)$r) # TRUE within floating-point tolerance
```

### Constraints

| Constraint | Detail |
|---|---|
| **Engine** | `"pca"` and `"efa"` only — `"esem"` errors (lavaan needs raw data) |
| **`n_obs`** | Required for `"efa"`; optional for `"pca"` (stored as `NA`) |
| **`cor` argument** | Ignored (basis is fixed); warns if set explicitly |
| **`missing` argument** | Ignored; warns if set explicitly |
| **Factor scores** | `keep_scores = TRUE`, `augment()`, `tidy(what = "scores")` all error |
| **`$cor` field** | Stored as `NA`; shown as `"(user-supplied matrix)"` in print |

### suggest_k() with a correlation matrix

```{r cor-matrix-suggest, eval = FALSE}
sk_R <- suggest_k(R, n_obs = 875L)
# CD is skipped (resampling requires raw item distributions)
# PA, MAP, and VSS run normally
```

## Performance with many items (ESEM)

Bass-ackwards analyses often involve large item pools (sometimes hundreds of
items), and the ESEM engine is the most expensive because it fits a separate
`lavaan` model at every level. Two things keep this manageable:

**Sample statistics are computed once.** For ordinal data (`cor =
"polychoric"`, WLSMV), lavaan must estimate item thresholds, the polychoric
correlation matrix, and an asymptotic weight matrix from the raw data. These
depend only on the data — not on the number of factors — so `ackwards()`
computes them once at the first level and reuses them for every deeper level.
The solutions are identical; only the redundant recomputation is removed. This
is the single biggest saving at large item counts, and it happens
automatically.

**Levels can be fit in parallel.** The per-level model fits are independent.
If you install the `future.apply` package, `ackwards()` dispatches them through
the [future](https://future.futureverse.org) framework. The default plan is
sequential (no change in behaviour); to parallelize, set a plan once before your
call:

```{r esem-parallel, eval = FALSE}
# install.packages(c("future", "future.apply"))
future::plan(future::multisession, workers = 4) # multicore on Linux/macOS

x <- ackwards(items, k_max = 8, engine = "esem", cor = "polychoric", seed = 1)

future::plan(future::sequential) # restore when done
```

Parallelism pays off when the per-level fits are genuinely heavy (large `p`,
several levels); for small problems the worker startup cost can outweigh the
gain. Results are reproducible across plans when you pass `seed`. PCA and EFA
already compute their correlation matrix once and do not need this.

If you only need the hierarchy (loadings and edges) and not ESEM's rotation-aware
standard errors and per-level fit indices, `engine = "efa"` with `cor =
"polychoric"` computes the polychoric matrix once and runs `psych::fa()` at each
level — substantially cheaper than k WLSMV fits — and recovers the same
structure.

## References

Goldberg, L. R. (2006). Doing it all bass-ackwards. *Journal of Research in
Personality*, *40*(4), 347–358.

Waller, N. G. (2007). A general method for computing hierarchical component
structures by Bass-Ackward factor analysis. *Journal of Research in
Personality*, *41*(4), 745–752.

Asparouhov, T., & Muthén, B. (2009). Exploratory structural equation modeling.
*Structural Equation Modeling*, *16*(3), 397–438.
